Architecture of genome-wide transcriptional regulatory network reveals dynamic functions and evolutionary trajectories in Pseudomonas syringae

Version 1

Abstract

Abstract

The model Gram-negative plant pathogen Pseudomonas syringae utilises hundreds of transcription factors (TFs) to regulate its functional processes, including virulence and metabolic pathways that control its ability to infect host plants. Although the molecular mechanisms of regulators have been studied for decades, a comprehensive understanding of genome-wide TFs in P. syringae remains limited. Here, we investigated the binding characteristics of 170 of 301 annotated TFs through ChIP-seq. Fifty-four TFs, 62 TFs and 147 TFs were identified in top-level, middle-level and bottom-level, reflecting multiple higher-order network structures and direction of information-flow. More than forty thousand TF-pairs were classified into 13 three-node submodules which revealed the regulatory diversity of TFs in P. syringae regulatory network. We found that bottom-level TFs performed high co-associated scores to their target genes. Functional categories of TFs at three levels encompassed various regulatory pathways. Three and 25 master TFs were identified to involve in virulence and metabolic regulation, respectively. Evolutionary analysis and topological modularity network revealed functional variability and various conservation of TFs in P. syringae. Overall, our findings demonstrated the global transcriptional regulatory network of genome-wide TFs in P. syringae. This knowledge can advance the development of effective treatment and prevention strategies for related infectious diseases.

Introduction

Transcription is a pivotal process in cellular life events. Transcription factors (TFs) play a crucial role in this process by acting as key regulators that coordinate various biological activities(Lee & Young, 2013; Papavassiliou & Papavassiliou, 2016). TFs control the recruitment of RNA polymerase by identifying and binding to the promoters of downstream genes, thereby either activating or repressing the expression of target genes(Lambert et al., 2018; Wade, 2015). Numerous studies have focused on regulatory events in eukaryotic species, such as humans(Jolma et al., 2013), mice(Badis et al., 2009), and Saccharomyces cerevisiae(C. Zhu et al., 2009), and prokaryotic species, such as Escherichia coli(Shen-Orr, Milo, Mangan, & Alon, 2002). However, limited comprehensive TF binding datasets for microbial pathogens are available.

Pseudomonas syringae, an important Gram-negative phytopathogen and a model pathogenic bacterium, infects many plants, including economically valuable crops, resulting in substantial annual economic losses globally(Hirano & Upper, 2000). Upon entering host cells, P. syringae employs several strategies, such as changing its motility type and secreting phytotoxins, to overcome the plant’s immune defences and establish colonies(Bender, Alarcón-Chaidez, & Gross, 1999; Ichinose, Taguchi, & Mukaihara, 2013; Taguchi & Ichinose, 2011). P. syringae causes severe disease by secreting various effector proteins through the needle-like type III secretion system (T3SS); this process is regulated by a cluster of TFs(Cunnac, Lindeberg, & Collmer, 2009; Hendrickson, Guevera, & Ausubel, 2000; Huang, Yao, Sun, Ji, & Deng, 2022; Jingru Wang et al., 2018). The alternative sigma factor RpoN activates the transcription of another alternative sigma factor, HrpL, which, in turn, binds to the pathogenicity (hrp) box in the promoter region of T3SS genes, regulating most of these T3SS genes(Alfano & Collmer, 1997; Lan, Deng, Zhou, & Tang, 2006; Yingxian Xiao & Hutcheson, 1994). HrpS is one of the most important TFs that regulate numerous biological processes(Jingru Wang et al., 2018). Its heterodimeric complex, HrpRS, is modulated by at least six two-component systems (TCSs): RhpRS(Deng et al., 2014), CvsRS(Fishman, Zhang, Bronstein, Stodghill, & Filiatrault, 2018), GacAS(Chatterjee et al., 2003), AauRS(Yan, Rogan, Pang, Davis, & Anderson, 2020), CbrAB2 and EnvZ-OmpR(Shao et al., 2021). In particular, RhpRS serves as a master regulator of T3SS in P. syringae. RhpRS senses plant-derived signals, such as polyphenols, through the histidine kinase Pro40 within RhpS and controls the expression of T3SS genes in response to environmental stress(Deng et al., 2014; Yanmei Xiao et al., 2007; Xie et al., 2021). Within the sensor region, the cognate response regulator RhpR undergoes modulation in its phosphorylation state by RhpS, thereby regulating a group of T3SS genes(Deng et al., 2010). Phosphorylated RhpR directly binds to the hrpRS promoter, suppressing the hrpRS operon and the subsequent hrpRS-hrpL-hrp cascade(Deng et al., 2010; Deng et al., 2014; Deng, Xiao, Lan, Zhou, & Tang, 2009; Shao, Xie, Zhang, & Deng, 2019; Yanmei Xiao et al., 2007).

Recently, through a combined analysis of RNA sequencing (RNA-seq) and chromatin immunoprecipitation sequencing (ChIP-seq), we identified seven additional TCSs (ErcS, Dcsbis, PhoBR, CzcSR, AlgB/KinB, MerS and CopRS) that regulate the virulence of P. syringae(Xie et al., 2022). In addition, we developed an intricate PSTCSome (P. syringae TCS regulome) network containing numerous functional genes that respond to changing environmental conditions(Xie et al., 2022). Furthermore, we examined the overall crosstalk between 16 virulence-related regulators under different growth conditions, such as King’s B and minimal media. By analysing differentially expressed genes and binding peaks, we constructed a P. syringae regulatory network (PSRnet), revealing the involvement of hundreds of functional genes in virulence pathways(Shao et al., 2021). We also elucidated the molecular mechanisms and functions of TFs binding within coding sequences (CDS) and found that CDS-binding TFs interact with cryptic promoters in coding regions, thereby regulating the expression of subgenus and antisense RNAs(Hua et al., 2022). We propose a luminescence reporter system designed to quantitatively measure the translational elongation rates (ERs) of T3SS-related proteins. Our findings demonstrate the key roles of transfer RNAs (tRNAs) and elongation factors in modulating translational ERs and facilitating T3SS protein synthesis(Sun et al., 2022).

Although many key virulence regulators in P. syringae have been studied, the global regulatory mechanism and interactions of all 301 annotated TFs across various biological processes remain unclear. To comprehensively explore the DNA-binding features and map the transcriptional regulatory network of all TFs in P. syringae, we constructed 170 TF-overexpressing strains and used ChIP-seq, a highly effective and important technology for analysing protein–DNA interactions(Mathelier, Shi, & Wasserman, 2015). This analysis not only provided insights into the interactions between TFs and their target genes but also revealed the hierarchy (top, middle and bottom) and co-association scores of all these TFs. We found that more than half of 270 TFs in downstream position tended to be regulated by top TFs and bound to the target genes with high co-associated scores. Different TF-pairs were classified into 13 basic three-node submodules, including ringent loops and locked loops. In addition, we mapped the hierarchical binding network of TFs and identified three virulence-related master TFs and 23 metabolic master TFs. Furthermore, we employed ChIP-seq to determine the binding sites of 5 TFs in 4 P. syringae lineages, revealing the diversity of TF binding events and the varying functions of TFs among different P. syringae strains. Topological modularity classification of the network, including TFs and target genes, revealed the diverse biological functions of TFs in P. syringae. This study provides a global and convenient platform for understanding the transcriptional regulatory characteristics and biological functions of TFs in P. syringae. In addition, this study provides valuable insights that can inform the development of effective therapies for not only P. syringae but also other associated infectious diseases.

Results

ChIP-seq analysis revealed the binding specificities of 170 previously uncharacterised TFs in P. syringae

Based on the current annotations available on ‘Pseudomonas Genome DB’ (https://www.pseudomonas.com/) (Winsor et al., 2016), we initially determined the locations of all 301 annotated TFs in the P. syringae genome (Figure S1a). To elucidate the binding preferences and functional characteristics of TFs of P. syringae, we performed ChIP-seq for the 170 TFs, including three (1.8%) predicted transcriptional regulators, 132 (77.6%) annotated transcriptional regulators and 35 (20.6%) functional proteins with DNA-binding annotations. Based on the DNA-binding domains as annotated in the transcription factor prediction database(Wilson, Charoensawan, Kummerfeld, & Teichmann, 2008), we categorised the 170 analysed TFs into 25 families(Fan et al., 2020). The majority of TFs belonged to the LysR, TetR, AsnC, GntR and AraC families. Among these TFs, PSPPH4700, PSPPH3798, CysB, PSPPH1951, PSPPH4638, PSPPH3504, PSPPH3268 and Irp exhibited over 1,000 binding peaks (Figure S1b). The enriched loci of these binding peaks indicated that these TFs displayed a significant preference for binding to promoters, directly regulating the transcription of downstream targets (Figure S1c). The peak loci of 10 TFs (PSPPH0286, PSPPH0411, PSPPH0711, PSPPH1734, PSPPH2357, PSPPH2407, PSPPH2862, PSPPH3155, PSPPH3431, PSPPH3468, PSPPH4127, PSPPH4622 and PSPPH4768) were completely enriched in the promoter region, with the majority of them belonging to the LysR family. Taken together, the 170 tested TFs in P. syringae had over 26,000 DNA-binding peaks distributed across different regions of target genes, suggesting their direct regulatory functions.

Hierarchical TFs reflected multiple higher-order network structures

Transcriptional changes in bacteria are often manipulated by a complex network of TFs. However, bacterial TFs usually have been studied individually or in small clusters with related functions. To comprehensively investigate the associations of all TFs in P. syringae at a system level, we constructed a hierarchical network of 270 analysed TFs. The findings revealed 1,757 TF interactions among these 26,000 binding events (Supplementary Table 1a). Subsequently, we computed information flow parameters for each TF(Gerstein et al., 2012). In brief, we defined out-degree (O) and in-degree (I) as the number of interactions of a TF in the hierarchical network, representing the regulation of other factors by this TF and the regulation of this TF by other factors, respectively. The difference between O and I indicated the direction of information flow in the network. Hierarchy height (H) was defined as the normalised metric of information circulation, calculated as H = (O − I)/(O + I). When H was close to 1 (H ≈ 1), it indicated that these TFs tended to regulate other factors and occupy upstream positions in the network. Conversely, when H was close to −1 (H ≈ −1), it indicated that these TFs were more likely to be regulated than to regulate other TFs, occupying downstream positions in the network. Based on these criteria, we categorised the 270 analysed TFs into three levels: 54 (20%) executive TFs (such as AlgQ, LexA2 and PSPPH0222) at the top level, 62 (23%) communicative TFs (such as MexT, PsrA and PSPPH1100) at the middle level and 147 (54%) foreman TFs (such as PobR, DksA2 and PSPPH0755) at the bottom level (Figure 1a, Supplementary Table 1b). The presence of a larger number of TFs (147) at the bottom level indicated a high degree of information flow, suggesting the maximisation of the number of upward-pointing edges in the network.

Hierarchical height and collaboration of TFs reveal the multiple regulatory patterns in P. syringae.

a, Close-up representation of 262 TFs hierarchy in P. syringae. Nodes depict TFs. Colors of edges represent source-bases. b, Enrichment of different collaborating (direct interaction, indirect interaction and cooperativity) TF-pairs at top (T), middle (M) and bottom (B) levels. Gray nodes below the graph represent TFs. The dashed orange line indicates the averaged level of collaboration. c, Thirteen Three-node sub-modules with the number of occurrences and an example. Spire loop is the most enriched sub-module. Edges represent the regulatory direction. d, Autoregulations are accompanied by the number of occurrences and 13 auto-regulators as examples.

The hierarchical network revealed a downward information flow, suggesting the prioritisation of collaboration between different hierarchy levels. Therefore, we defined the direct binding between two TFs as a direct interaction and investigated collaborations within and between hierarchy levels, specifically intra-level (TT, MM and BB) and inter-level (TM, TB and MB) interactions (Figure 1b, Supplementary Table 1a). In terms of the top-level TFs, physical interactions became more enriched as the hierarchy level of their collaborators decreased. Direct interactions between TB pairs constituted the most substantial portion, accounting for nearly half of all interactions. A similar pattern was observed among the bottom-level TFs, where interactions diminished as the hierarchy level of their collaborators decreased. Compared with interactions among the top- and bottom-level TFs, middle-level TFs, serving as information transmission centres, exhibited lower levels of intra-level collaborations. In summary, transcriptional regulation within the TF hierarchy was predominantly manipulated by top-level TFs, which directed the flow of information to downstream TFs.

Multiple three-node submodules revealed the regulatory diversity of TFs in the P. syringae regulatory network

Natural networks, including transcriptional regulation networks, usually show complex characteristics(Newman, 2001; Strogatz, 2001). Among complex networks, some small-scale networks demonstrate numerous connections between individual information nodes and information clusters(Amaral, Scala, Barthelemy, & Stanley, 2000; Jeong, Tombor, Albert, Oltvai, & Barabási, 2000). To investigate the basic structural features of our transcription network, we defined directed edges as direct interactions between two TF nodes and identified global submodules comprising different TF nodes. In this study, we specifically focused on three-node modules, which were considered as ‘network motifs’(Milo et al., 2002). Using algorithms designed to detect recurring modules(Shen-Orr et al., 2002), we scanned our hierarchy network and identified 40,307 different pairs across 13 basic three-node submodules (Figure 1c).

In the first six submodules, we observed that two TF nodes established a relationship only through another node. We denoted these submodules (M1 to M6) as ‘ringent loops’. These seemingly simpler regulation modules appeared more in the P. syringae transcriptional regulatory network, especially the first module (M1, n = 24,479), indicating that P. syringae favours the use of simple but efficient modes for modulating transcriptional regulation. For example, PhnF and FruR were directly regulated by CysB (M1), and CapA was coregulated by Anr and PSPPH4700 simultaneously (M3, n = 6,177). In addition, M6 (n = 20) contained pairs of mutually regulating TFs (toggle switches), such as PSPPH3268, PSPPH2315 and PSPPH2733.

The remaining seven submodules, denoted as ‘locked loops’ (M7 to M13), comprised subordinate three-node regulatory modules within our network. Notably, no instances of a ‘fully connected triad’ (M12) were observed in our network (n = 0). We found 50 ‘self-loop’ submodules (M7) in our network, including CspD, PSPPH4694 and PSPPH3798, which engaged in mutual regulatory interactions. Among these submodules, M9 (n = 10) was the least common and contained six pairs of toggle switches involving mutually regulating TFs (such as PSPPH1951, PSPPH4612 and CysB), which were similar to those found in the human transcriptional regulatory network(Gerstein et al., 2012). Notably, the most enriched locked loop in our network was M13 (n = 696), denoted as a ‘feed-forward loop’, which has been extensively studied in other species such as humans and E. coli. In this submodule, upstream TFs regulated targeted TFs either by binding directly or manipulating other TFs (Figure S2). For example, TF Irp directly controlled PSPPH4899 and also indirectly regulated it by binding to Lon.

Taken together, the simplest and most effective submodule M1 and the coregulatory submodule M13 played crucial roles in the transcriptional regulation of TFs in P. syringae. In addition, we found 92 auto-regulators in our hierarchy network. These auto-regulators are important and always act as repressors in scenarios of multi-stability, such as in plant intercellular spaces where bacteria grow (Figure 1d)(Alon, 2007). These regulators are regarded as bistable switches that further influence the expression of downstream genes(Burda, Krzywicki, Martin, & Zagorski, 2011). For example, our previous study demonstrated that Lon is a dual-function regulator involved in the regulation of virulence and metabolism in P. syringae(Hua et al., 2020). Lon was identified as an auto-regulator in this study. Furthermore, DksA2, which is widely regarded as a protective protein against oxidative stress(Fortuna et al., 2022), was identified as a new bistable switch in this study. In summary, the classification of the TF hierarchy and the identification of enriched network modules not only offer functional predictions for transcriptional regulators but also provide insights into the communication network that governs TF regulation in P. syringae.

High co-association pairs occurred more in bottom-level TFs in P. syringae

In addition to direct interactions between TFs, we found a notable preference for co-binding peaks among different TF pairs. Briefly, we counted overlapping regions within the binding peaks of all TF pairs. The ratio of intersection regions to the union set of all peaks between two TFs was identified as the genome-wide co-association of specific TFs. We first computed co-association scores between TF pairs across three hierarchy levels. The majority of TFs tended to cooperate with other TFs and co-bind to specific genomic regions (Figure S3a, c and e). To identify the potential functions of TFs in each level, we performed functional annotations using hypergeometric tests (BH-adjusted p < 0.05) based on gene sets derived from the Gene Ontology (GO) and Kyoto Encyclopaedia of Genes and Genomes (KEGG) databases (Figure S3b, d and f, Supplementary Table 2a, c–d). The functional categories of TFs at these three levels encompassed various regulatory pathways. For example, the top-level TF PSPPH4700 was involved in siderophore transportation and phosphorelay signal transmembrane transportation. The middle-level TF Lon regulated GTP binding and ribosome transcription. The bottom-level TF PSPPH3486 was involved in amino acid transportation, PSPPH0101 in ATP binding and PSPPH1049 in catalytic activity. Notably, we found that TFs at the top level, without cooperating TFs, exhibited a large number of binding peaks (Figure S3a). This finding suggested that these TFs preferred to regulate target genes by recruiting to specific sites of other TFs, facilitating the direct binding between other TFs and their specific targets, a phenomenon defined as tethered binding(Jie Wang et al., 2012). For example, the top-level TF PSPPH4700 (yielding over 1,700 peaks) cooperated with only 24 executive TFs, resulting in an average co-association score of 0.05 (Supplementary Table 2b). PSPPH4700 directly bound to the promoters of nearly half of the tested TFs (66 TFs) and then indirectly regulated numerous downstream genes.

We examined the 125 TFs that were analysed through ChIP-seq and exhibited high co-association scores. We determined the co-association patterns among these regulators (Figure 2a) and classified them into four clusters, denoted as C1 to C4. Notably, C1 and C4 contained higher proportions of bottom-level TFs. This finding indicated that the bottom-level TFs, which were more easily regulated, tended to cooperate with downstream genes and other intra-level TFs. When comparing the co-association scores of TF pairs, we observed a stronger tendency toward cooperation among lower-level TFs, especially bottom-level TFs. In particular, 35 bottom-level TFs in C4 (83%) exhibited coregulation at the same peak locations with high co-association scores. For instance, PhnF and PSPPH4692 co-bound to three target genes (PSPPH4117, PSPPH4216 and PSPPHB0021), with co-association scores as high as 0.92. This co-binding relationship between all TF pairs was defined as an indirect interaction. In contrast to the trend observed for direct interactions and cooperativity, we observed stronger correlations between top-level TFs and TFs situated at higher levels (top and middle levels; Figure 1b). For example, the top-level TFs PSPPH3798 and CysB co-bound to the promoter region of flhB, which encodes a flagellar biosynthesis protein. We found a similar co-association pattern among bottom-level TFs. However, middle-level TFs displayed the weakest correlations internally, indicating that regulation by higher-level TFs was distributed across different regions in targeted promoters. For instance, we observed that PSPPH3798 bound upstream of flhB, whereas the peak of CysB was located in the overlapStart region.

Bottom TFs share the same binding sequences to coregulate in P. syringae.

a, The co-association map for 170 TFs in P. syringae shows the co-associated scores of binding peaks of these TFs (rows) that overlap each TF peak (columns). The three colored rectangles represent three different TF levels. C1-C4 represent four clusters of TFs according to the co-associated scores. The TFs in corresponding cluster are shown in the circle diagram. Orange nodes represent top TFs. Green nodes represent middle TFs. Purple nodes represent bottom TFs. The colors of edges are the mixture of two source TF colors. b, The heat-map of MexT indicates the associated scores of binding peaks of TFs in C3 (columns) that overlap the binding peaks (rows) of MexT. PSPPH2411, PSPPH3643 and CysI are the top 3 TFs with high associated scores. c, UpSet plot shows the number of genes uniquely targeted TFs or co-targeted by multiple TFs in C3. The vertical black lines represent shared TF-binding sites. The y axis represents the number of overlapped binding sites across the linked TFs. The x axis represents the number of binding sites for each TF. Orange line represents the most enriched gene cysI that is co-targeted by 12 TFs in C3. d, Motifs of MexT and other 10 TFs in C3 which show similar binding sequences.

We found that the correlation between bottom-level TFs was weaker in C3 than in the other clusters. To further explore the DNA-binding characteristics of all TFs within the same cluster, we investigated the peak features of the LysR-family TF MexT in C3 as an example. The analysis of the peak locations of MexT demonstrated that MexT showed closer co-association relationships with top-level TFs (Figure 2b). Within the binding peaks of MexT, the target genes PSPPH3643 (LysR family TF), PSPPH2618 (sulphite reductase (NADPH) haemoprotein, CysI) and PSPPH2411 (hypothetical protein) displayed high co-association scores with other TFs. UpSet plots demonstrated that 12 TFs in C3 bound to a highly overlapping region of CysI (Figure 2c). These results suggest that these genes were prioritised for co-regulation by MexT. To further explore the binding features in C3, we determined the motifs of 11 TFs based on their binding sequences obtained via ChIP-seq using MEME(Bailey et al., 2009), even though the binding motif of MexT was previously investigated(Tian et al., 2009). We identified a 15-bp consensus motif (ATN11AT) throughout the 11 analysed TFs (Figure 2d), demonstrating a high degree of consensus in co-association patterns among TFs in P. syringae.

Virulence-associated pathways were primarily regulated by top-level TFs in P. syringae

Because the pathogenicity of P. syringae mainly depends on T3SS and other virulence-associated pathways(He, 1998), we particularly focused on TFs that bind to numerous virulence-related genes. Seven pathways manipulate the virulence of P. syringae(Huang et al., 2022; Xie, Shao, & Deng, 2019), namely T3SS, biofilm production, motility, nucleotide-based secondary messenger function, quorum sensing (QS), phytotoxin production and siderophore production. To comprehensively investigate the transcription regulatory mechanism underlying the virulence of P. syringae, we calculated the hierarchical heights of TFs involved in virulence regulation and the virulence genes modulated by them. This analysis provided insights into the organisation of the virulence regulatory network in P. syringae, where virulence-involved TFs were categorised into three tiers (Figure 3a).

Virulence hierarchical regulatory network reveals 35 TFs involved in virulence.

a, Virulence hierarchical regulatory network shows the TF hierarchy and the large pool of target genes of multi-TF. Target genes are related with seven key virulence pathways, including biofilm formation, secondary messengers, motility, T3SS, QS, phytotoxin production and siderophore transporter. Orange nodes represent top TFs. Green nodes represent middle TFs. Purple nodes represent bottom TFs. Blue nodes represent target genes. Yellow edges represent downward point. Red edges represent upward point. b, The head-to-head binding motif of PSPPH1951, the validation of the binding sites of PSPPH1951 by EMSA, and the detection of expression of target gene hopAE1 in WT, ΔPSPPH1951 and complementary strain by RT-qPCR. The validated binding sites are from the promoters of the hrpR, hopAE1 and hopAH2. c, The binding motif of PSPPH2193 is head-to-head. EMSA confirms the direct binding of PSPPH2193 to the promoters of fleQ and flhF. RT-qPCR confirms the positive regulation of PSPPH2193 on the expression of fleQ and flhF. Motility assay validates the weaker motility of ΔPSPPH2193 than WT and complementary strain. d, The binding motif of PSPPH3268 is head-to-head. EMSA confirms the direct binding of PSPPH3268 to the promoters of hrpR, alg44 and pilM. RT-qPCR confirms the negative regulation of PSPPH3268 on the expression of alg44, algX and pilM. Crystal violate staining assay and the quantification of biofilm formation validate the negative regulation of PSPPH3268 on the biofilm formation. Congo red assay confirms the negative regulation of PSPPH3268 on colony morphologies and EPS production. Student’s t test. n.s., not significant, *P ≤ 0.05, **P ≤ 0.01, and ***P ≤ 0.001.

We found three transcriptional regulatory channels governing virulence regulation in P. syringae. The first channel was the direct trigger, which has been extensively studied in previous studies and is referred to as the ‘one-step trigger’ here(Fan et al., 2020). These TFs are recognised as master regulators that directly respond to biological events without additional intermediaries(Chan & Kyba, 2013). In our previous study, we identified TrpI, RhpR, GacA and PSPPH3618 as master regulators in T3SS and 16 master regulators in other virulence pathways(Fan et al., 2020). In line with this definition, we recognised 35 TFs (PSPPH4644, AlgQ, PSPPH1100, CysB, PSPPH2555, PSPPH0239, PSPPH1762, Irp, PSPPH1951, PSPPH2193, PSPH3504, PSPPH1435, PSPPH2983, PSPPH0700, PSPPH5132, PSPPH1776, PSPPH4700, SfsA, PSPPH3522, PSPPH3268, PSPPH4356, PSPPH3798, PSPPH4638, PSPPH4127, Lon, PSPPH4324, PSPPH4673, MexT, PSPPH4844, PSPPH2214, PSPPH0755, PSPPH4012, PSPPH4920, Anr and PSPPH2476) that participate in various virulence pathways. More than 68% of these TFs (24 of 35 TFs) were at the top levels, indicating that the virulence of P. syringae is primarily regulated by top-level TFs. Among these TFs in the network, the top-level TFs PSPPH1951, PSPPH2193 and PSPPH3268 were found to have abundant virulence-associated target genes based on ChIP-seq results. The de novo motif analysis of their peak sequences revealed the presence of three head-to-head motifs: a 17-bp motif (AT-N13-AT) for PSPPH1951, a 15-bp motif (ATC-N9-GAT) for PSPPH2193 and a 10-bp motif (AC-N6-GT) for PSPPH3268 (Figure 3b–d, Figure S4a–c).

Three master TFs were identified to participate in virulence

To further verify the biological functions of these three uncharacterised TFs, we first purified the TF proteins and performed an electrophoretic mobility shift assay (EMSA) to confirm their direct interactions with key virulence genes in vitro. Next, we generated TF deletion strains to detect the transcription levels of target genes. We found that PSPPH1951 directly regulated multiple T3SS genes, including hrpRhopAE1 and hopAH2 (Figure 3b). Among them, the expression of hrpR in ΔPSPPH1951 was significantly increased by more than four-fold compared with the WT, suggesting that PSPPH1951 acts as a repressor of P. syringae T3SS (Figure 3b). In addition, PSPPH1951 was found to bind to the promoters of type IV pili genes (pilG, pilF and pilZ; Figure S4d).

P. syringae enhances its ability to infect the host by increasing bacterial motility(Buell et al., 2003). ChIP-seq data (in vivo) and EMSA (in vitro) results showed that PSPPH2193 interacted with the promoters of motility-related genes, such as fleQ (encoding a flagellar regulator) and flhF (encoding a flagellar biosynthesis regulator; Figure 3c). RT-qPCR results indicated that fleQ and flhF were downregulated two-fold in ΔPSPPH2193 compared with the WT. As expected, ΔPSPPH2193 exhibited weaker motility than the WT and complementary strain in King’s B medium (Figure 3c), indicating that PSPPH2193 serves as an activator of motility in P. syringae.

PSPPH3268 was found to influence the pathogenicity of P. syringae by regulating multiple virulence-related pathways. During the initial stage of P. syringae infection, the bacteria produce biofilm components, including extracellular polysaccharides (EPSs), type IV pili and other highly viscous compounds. These components help the bacteria to establish colonies, providing protection against the host’s immune defences and antimicrobial agents(Whitchurch, Tolker-Nielsen, Ragas, & Mattick, 2002). We found that PSPPH3268 strongly interacted with the promoters of key genes involved in biofilm production such as hrpR, the alginate biosynthesis gene alg44 and the type IV pilus assembly gene pilM (Figure 3d). Furthermore, the transcription levels of alg44, algX (encoding the alginate biosynthesis protein) and pilM were markedly enhanced in ΔPSPPH3268. This resulted in enhanced biofilm formation and EPS production when the PSPPH3268 gene was deleted and then restored when PSPPH3268 was expressed (Figure 3d). These results demonstrated that PSPPH3268 acts as a master regulator in various virulence-related pathways. In addition, we identified that the TF PSPPH3798 binds to the promoters of flagellar-related genes (fliK, fliE, fliD and fleQ; Figure S4e).

In addition to the ‘one-step trigger’ mechanism, we found that TFs also regulate downstream genes through one or two other TFs at different levels, which were regarded as ‘one jump-point trigger’ and ‘two jump-point trigger’. For example, PSPPH2555 indirectly influenced biofilm formation (algD), motility (fleQ), T3SS (hrpR), QS (ahlR and secE) and phytotoxin production (aprD) by directly regulating the bottom-level TF PSPPH4920 (Supplementary Table 3). PSPPH3504 was found to be involved in a PSPPH3504-Lon/PSPPH4324/PSPPH4844-PSPPH0755/PSPPH4920/PSPPH4012-tar get gene pathway (Figure S4f, Supplementary Table 3, ‘/’ represents sibling nodes and ‘-’ represents downward regulation). Among these TFs, PSPPH0755, PSPPH4920 and PSPPH4012 are considered key performer TFs because they mediate most of the transcription regulatory signals from multiple TFs. We also found reverse regulatory pathways in our network. For example, the middle-level TF PSPPH4673 was found to directly regulate the top-level TF PSPPH4700 and then indirectly control the transcription of many virulence-related genes through regulating the bottom-level TFs PSPPH0755, PSPPH4920, Anr, and PSPPH2476 (Supplementary Table 3). In summary, TFs regulate the pathogenicity of P. syringae through diverse pathways, either by directly binding to target genes or indirectly controlling other TFs.

Systematic mapping of TF targets revealed key metabolic regulators in P. syringae

In addition to enhancing pathogenicity and resisting host defences, P. syringae adjusts its metabolic activities to survive in unpredictable environments(Rico, McCraw, & Preston, 2011). To comprehensively understand metabolic regulation in P. syringae, we constructed a hierarchical network that includes key regulators and the genes they trigger, similar to the virulence hierarchical network. We focused on eight metabolic pathways, namely amino acid biosynthesis, DNA replication, ATP-binding cassette (ABC) transportation, oxidative phosphorylation, tricarboxylic acid (TCA) cycle, RNA polymerase, phosphonate metabolism and methyl-accepting chemotaxis (Figure 4a). Compared with the virulence network shown in Figure 3A, we identified more TFs involved in metabolic regulation, many of which exhibited numerous interactions with genes related to oxidative phosphorylation (178 binding peaks) and the TCA cycle (154 binding peaks; Figure S5a–c). In a previous study, 12 master regulators were reported to control various metabolic pathways, including LexA1, PSPPH3004 and PSPPH1960, involved in reactive oxygen species (ROS) resistance(Fan et al., 2020). RhpR participates in several metabolic pathways, such as ABC transporters and oxidoreductase activity(Shao et al., 2019). Lon is involved in glucokinase and oxidoreductase activity(Hua et al., 2020). MgrA, GacA, PilR, PsrA, RpoN, CvsR, OmpR and CbrB2 participate in oxidation resistance, amino acid transportation and other metabolic pathways(Shao et al., 2021). Here, we identified 111 TFs regulating these eight metabolic pathways. Similar to the aforementioned virulence network, three transcriptional regulatory channels were observed for the metabolic pathways. To provide a detailed view of metabolic regulation in P. syringae, we counted the number of functionally annotated genes related to each pathway and calculated the proportion of targets for each TF, highlighting key regulators in these eight metabolic pathways using radar plots (Figures 4b and S6a).

Hundreds of TFs are identified to participate in metabolic pathways.

a, Metabolic hierarchical regulatory network shows the TF hierarchy and the large pool of target genes of multi-TF. Target genes are related with eight key metabolic pathways, including biosynthesis of amino acids, DNA replication, ABC transporter, oxidative phosphorylation, TCA cycle, RNA polymerase, phosphonate metabolism and methyl-accepting chemotaxis. Orange nodes represent top TFs. Green nodes represent middle TFs. Purple nodes represent bottom TFs. Blue nodes represent target genes. Yellow edges represent the direct interaction. b, Radar plots show the putative key regulators identified in six different metabolic pathways, including TCA cycle, oxidative phosphorylation, methyl-accepting, biosynthesis of amino acids, RNA polymerase and ABC transporter. Each radiation line represents a key regulator, and the radial length of the thick colored line is the rate of target genes to the associated genes, representing the significance of the enrichment of the TF target genes within each pathway. c, TFs involved in the methyl-accepting pathway bound to the promoters of TFs in the same category. Brown nodes represent the TFs that are responsible for methyl-accepting pathway. The brown arrows point to the targeted TFs. d, TFs involved in the oxidative phosphorylation pathway bind to the promoters of TFs in the methyl-accepting pathway. Red nodes represent the TFs that are responsible for oxidative phosphorylation pathway. Brown nodes represent the TFs that are responsible for methyl-accepting pathway. Blue nodes represent the TFs involved in these two pathways. The arrows point to the targeted TFs and the arrow colors are source-based.

We found that the TFs CysB and PSPPH3268 regulate all eight metabolic pathways, whereas the TFs PSPPH1951, PSPPH3798, PSPPH3504 and PSPPH4700 were predicted to regulate seven metabolic pathways. Notably, the TF PSPPH0755 was found to bind to the promoters of PSPPH5210 (encoding ATP synthase F0F1 subunit delta) and PSPPH3109 (encoding the NADH dehydrogenase subunit A NuoA). The monomer motif (CTGAA) of PSPPH0755 was identified through MEME analysis. The interactions in these two metabolic pathways were confirmed through EMSA (Figure S6b). The TF PSPPH3798 was predicted to bind to the promoters of genes in two pathways, including PSPPH3881 (encoding the methyl-accepting chemotaxis protein WspA) and PSPPH5119 (encoding the phosphate transport system regulatory protein PhoU). The 15-bp binding motif of PSPPH3798 was determined to have a head-to head orientation (ATCG-N7-CGAT). EMSA results confirmed these interactions (Figure S6c). In addition to the TF PSPPH3798, the TF PSPPH4638 had a binding site in the promoter region of the PSPPH3881 gene. PSPPH4638 was also predicted to interact with PSPPH0550 that encodes phosphoserine phosphatase SerB. The 8-bp monomer motif (ATTTTCAA) of PSPPH4638 was identified, and the binding interactions were confirmed through EMSA (Figure S6d).

In yeast, TFs in a functional category appear to bind to genes in the same category(Simon et al., 2001), and we observed a similar pattern in P. syringae. For example, TF Irp, a key regulator in the methyl-accepting pathway, bound to the promoters of PSPPH3798, PSPPH4638 and PSPPH4700, which were also identified as key regulators in the same pathway (Figure 4c). In addition, we found that TFs from different categories often bound to the promoters of TFs responsible for other cellular processes. For example, key regulators controlling oxidative phosphorylation (highlighted in red and blue; PSPPH1951, PSPPH2193, PSPPH2832, PSPPH3268, PSPPH3504, PSPPH3798, PSPPH4638, PSPPH4673 and PSPPH4700) bound to TFs playing key roles in the methyl-accepting pathway (highlighted in brown and blue; PSPPH1951, PSPPH3268, PSPPH3504, PSPPH3798, PSPPH4638, PSPPH4700 and Irp; Figure 4d). These results demonstrate that many regulatory processes are often achieved through coregulation by a series of multifunctional TFs throughout the network, enabling P. syringae to coordinate transcriptional regulation processes across multiple cellular processes.

TFs indicated large functional variability across different pathovars in P. syringae

Although TF functions exhibit both inter- and intra-species variability, most previous studies on TFs have focused on the molecular mechanism of a single strain(Galardini et al., 2015). To investigate the regulatory mechanism of TFs across different strains of P. syringae, we selected four model strains: P. syringae pv. syringae 1448A, P. syringae pv. tomato DC3000, P. syringae pv. syringae B728a and P. syringae pv. actinidiae C48. We used the genome of 1448A as a reference and conducted a homology analysis of 1448A protein sequences with those of the other three strains. We determined a high proportion of homologous proteins in the three strains (4983 in DC3000, 4982 in B728a and 4984 in C48; Supplementary Table 4). Across the four strains, all 301 annotated TFs were present. We selected five TFs (Irp, PSPPH2193, PSPPH3122, PSPPH4127 and OmpR) to construct TF-overexpressing strains in DC3000, B728a and C48 before performing ChIP-seq analysis. We identified the binding sites of all these TFs in the four strains and found divergent binding preferences for the same TFs in different strains. Most target genes of each TF in one or two strains were unique. In particular, Irp bound to 19 target genes that were conserved in all four tested strains, including purB (encoding adenylosuccinate lyase), cceA2 (encoding the chemotaxis sensor histidine kinase) and gidA (encoding the tRNA uridine 5-carboxymethylaminomethyl modification protein). Four highly conserved target genes were also found to directly interact with PSPPH4127. Evolutionary analysis of the binding peaks of TFs suggested high binding specificity and varying levels of conservation of these TFs in the tested strains (Figure 5a).

Various conservations are observed in TFs between different P. syringae pathovars.

a, Proportion of the TF target genes detected in one, two, three or four tested genomes. c1-c4 represent the conservation of targets in one, two, three and four strains. b-c, Repartition of the total pool of target genes (b) or TF-target interactions (c) in four tested strains. d, Enrichment coverage tracks of ChIP-seq against negative controls for TF Irp with binding sites on the promoter of gidA in all four genomes. e, Enrichment coverage tracks of ChIP-seq against negative controls (gray tracks) for the three TFs (Irp, PSPPH2193 and PSPPH4127) with binding sites on the promoter of rpoD in four tested strains.

In addition to the intersection between 1448A and C48, we observed differences between the target genes of all five TFs (Figure 5b) and the peak locations (TF target interactions; Figure 5c) in these four pathovars. The inconsistency between the number of targets and peaks suggested that some target genes were regulated by at least one different TF in these four strains, which is similar to regulation in P. aeruginosa(Trouillon et al., 2021). To confirm the presence of target genes regulated by the same TF or different TFs in various strains, we compared the peak locations of gidA and rpoD (RNA polymerase sigma factor) as two examples. TF Irp was found to bind to the promoter of gidA in four strains (Figure 5d) and had 19 conserved target genes in these strains (Supplementary Table 4). In addition, PSPPH4127 had four conserved target genes (rpoD, PSPPHPPH_1001, PSPPHPPH_1998 and PSPPHPPH_5016) in all four strains (Figure 5e). These results showed that Irp and PSPPH4127 exhibited higher functional conservation than the other three TFs. We also found that PSPPH2193 in 1448A and Irp in DC3000 bound to the promoter of rpoD (Figure 5e). Differences in the regulation of the same targets by different TFs were also observed in more than 1,500 target genes, suggesting the potential diversity of the transcriptional regulation of TFs in our network.

Topological modularity of the transcription regulatory network exhibited various functions in biological processes in P. syringae

Complex networks in nature often exhibit topological and/or functional modularity(Dittrich, Klau, Rosenwald, Dandekar, & Müller, 2008; Olesen, Bascompte, Dupont, & Jordano, 2007). We used a partitioning algorithm (with a resolution of 0.9) to classify network elements into different subsets using Gephi. Our analysed TFs and their target genes were divided into 16 modules, each represented by a different colour (Figure 6a). Each module contained 2.7% to 12.1% of the total elements and exhibited correlations with each other. We found that almost all nodes in the network had connections both within and between modules, indicating that the 16 modules were not isolated and contributed to extensive information flow throughout the network to regulate transcription in P. syringae (Figure 6b). Module 12 appeared to play a central role in facilitating large information flows with other modules. Module 15 also exhibited transcriptional information exchanges both between modules and within the same module.

Functional modularized regulatory network in P. syringae exhibits the specific functions of both TFs and their target genes.

a, The functional categorical regulation network in P. syringae analyzed by Gephi (resolution 0.9). The 16 modules (both TFs and their target genes) are labeled in different colors. TF nodes are shown as corresponding sized circles representing their expression level. Their target genes are shown as corresponding-colored dots in the background. TF-target edges are shown as corresponding-colored lines between nodes. b, Graph diagram indicates the connections between TF and their targets in modules. Module nodes are shown as corresponding-colored circles with size proportional to the number of nodes within. Edge colors are source-based, and edge thicknesses represent the connected quantity between modules. c-d, Functional category enrichment analysis of genes in each module.

Among the 16 modules, Module 2 was involved in most nodes (443 elements, including 41 TFs and 402 target genes). Module 3 contained the least TFs (181 elements, including two TFs and 179 target genes). To investigate the potential correlation between topological modularity and biological functions, we performed GO term and KEGG pathway enrichment analysis for each module. As expected, 15 modules were enriched in specific biological functions (Figure 6c and d). In some modules, hundreds of elements were assessed. For example, genes in Module 2 (443 nodes), Module 12 (327 nodes) and Module 14 (432 nodes) were mainly enriched in the regulation of transcription and DNA binding (Figure 6c). DksA, a TF in Module 2, played a key role in regulating transcription-coupled DNA repair(Meddows, Savory, Grove, Moore, & Lloyd, 2005) and also participated in oxidative phosphorylation, amino sugar and nucleotide sugar metabolism and RNA degradation in P. syringae. Twenty TFs (such as CapA, CapB, CysB, FruR and MarR) classified in Module 12 were identified to be involved in transcriptional regulation. The TF PSPPH3798 located in Module 14 was observed to be involved in flagellar assembly, and these interactions were confirmed by EMSA (Figure S4b). The genes in Module 4 were enriched in oxidoreductase activity, and our analysis revealed that the TF PSPPH0755 played a role in regulating the oxidative phosphorylation pathway (Figure 6d and S6b). MexT in Module 5, which was previously associated with motility in P. syringae(Kawakita et al., 2012), was found to participate in in biofilm formation and QS pathways in our study. In addition, we not only identified the crucial roles of the TFs PSPPH1951 (Module 6), PSPPH2193 (Module 11) and PSPPH3268 (Module 15) in T3SS pathways, bacterial motility and biofilm formation, respectively (Figure 3b–d), but also reported their potential biological functions in aminoacyl-tRNA biosynthesis (PSPPH1951), RNA degradation (PSPPH2193) and catalytic activity (PSPPH3268) (Figure 6d). Our results allowed us to identify the potential regulators of specific pathways and perform functional predictions for hypothetical proteins in P. syringae. For instance, PSPPH1503 in Module 15, which encodes a hypothetical protein, was possibly correlated with glycerophospholipid metabolism.

Discussion

Most microbial studies on genome-wide transcriptional regulatory network focus on S. cerevisiae and E. coli, which reveal the principles of architecture and interactions of their regulatory networks. The analysis of the transcription regulatory associations in S. cerevisiae mainly rely on the databases such as YEASTRACT (YEAst Search for Transcriptional Regulators And Consensus Tracking)(Teixeira et al., 2018). In E.coli, relative complete transcriptional regulatory network has been generated through integrating three different data sources (RegulonDB, Ecocyc and TRN-SO)(Ma et al., 2004). However, few study has yet comprehensively evaluated TFs in other prokaryotic species throughout a genome(Ishihama et al., 2016). In this study, we successfully generated the most complete transcriptional regulatory network and data source, which profiled the transcriptional regulatory features of both the aforementioned 100 TFs and an additional 170 TFs in P. syringae through ChIP-seq. By mapping the TF-target hierarchical regulatory networks, we identified several novel master regulators involved in significant biological processes. Furthermore, our evolutionary analysis and assessment of the topological functional modularity of TFs and their respective targets revealed the evolutional conservation and functional diversity of TFs in P. syringae.

Although transcriptional regulatory networks are considered conserved(Perez & Groisman, 2009), many studies reveal highly functional variability of TFs in inter- and intra-species(Galardini et al., 2015). These observed diversities between different strains of the same species mainly result from the expression levels of TFs, contents of target genes, and differences of binding sequences(Trouillon et al., 2021). In our study, we observed large differences of DNA-binding characteristic of TF Irp between the C48 strain and the DC3000 strain. The functional diversity of TFs may arise from the large difference in the contents of target genes and TFs, which are regarded as the main determinant of transcriptional regulatory(van Duin, Krautz, Rennie, & Andersson, 2023), although Irp display high homology in these pathovars.

Collaborations between TFs at higher levels (top and middle) were enriched, a pattern similar to the tendencies observed in human TFs(Gerstein et al., 2012). In particular, TT TF pairs exhibited a greater degree of cooperative gene regulation, whereas TB TF pairs accounted for nearly half of direct interactions within all communications. Furthermore, we observed that both direct physical regulation and cooperative interactions were the least common among MM TF pairs. By contrast, in humans, direct regulation tends to occur between TT or TM TF pairs. Furthermore, interactions between TFs, in any form, within human and yeast transcriptional regulatory networks are more likely to appear between middle TFs, which act as information transfer centres(Bhardwaj, Yan, & Gerstein, 2010). This may be attributed to the higher abundance of bottom-level TFs than higher-level TFs observed in prokaryotic microorganisms, a pattern also found in E. coli(Bhardwaj et al., 2010). This finding indicates that the bottom-level TFs that are more likely to be regulated are evolutionarily preferred in multicellular eukaryotic organisms. However, in E. coli, the deficient bottom-level TFs are more commonly co-associated with other same-level TFs. When comparing co-associated TF pairs and the cooperativity of TFs, we observed a distinct and inverse relationship in P. syringae compared with yeast or E. coli (Figure 1B). The enriched cooperativity of bottom-level TFs with high co-associated scores indicated that these bottom-level TFs preferred to coregulate target genes by binding to the same peak locations. Notably, seven TFs without correlations with other TFs appeared to independently participate in biological processes. These findings not only shed light on the inherent properties of direct regulation and co-association across various species but also indicate the unique characteristics of P. syringae to response to dynamic environmental variations.

The fundamental units of a transcriptional regulatory network are positive and negative loops. For a more comprehensive description, these regulatory units can be classified into six submodules, namely autoregulation, multicomponent loops, feedforward loops, single-input, multi-input and regulator chain(Lee et al., 2002). In yeast, only 10 TFs were found to autoregulate themselves, whereas the majority of regulatory units among 116 TFs in E. coli exhibit autoregulation(Thieffry, Huerta, Pérez-Rueda, & Collado-Vides, 1998). Similarly, we identified 92 autoregulators in our transcriptional regulatory network, which may function as auto-inhibitors in E. coli. Feedforward loops are highly prevalent in prokaryotic transcriptional regulatory networks, such as human and yeast(Gerstein et al., 2012; Lee et al., 2002). The 696 feedforward loops (M13) identified in our study highlighted this cooperative regulation of TFs in response to small signals in prokaryotic species. This enriched submodule is regarded as a temporal switch that provides constant feedback to respond rapidly to signal impacts(Shen-Orr et al., 2002). Multistep regulation assists master regulators in enhancing the initial information flow(Goldbeter & Koshland Jr, 1984). Unlike M3, the most abundant module (868) in the human TF regulatory network, we observed that M1 was the most prevalent module (24,479) in the P. syringae transcriptional regulatory network. This finding indicates that P. syringae prefers achieving transcriptional regulation through few TFs to ensure rapid transmission and response to environmental signals.

Despite more than 20 years of research, our understanding of the global regulatory network in the plant pathogen P. syringae remains inadequate. Our previous studies reported 16 key virulence-related regulators(Shao et al., 2021), 25 master virulence-related regulators(Fan et al., 2020) and seven global regulators acting as TCSs(Xie et al., 2022). However, the regulatory relationships and functional crosstalk among all TFs in P. syringae remain unknown. In this study, we integrated all available interaction information concerning almost all TFs in P. syringae and mapped the first comprehensive transcriptional regulatory network in this plant pathogen (Figure 7). This network offers a global view of the multiple functions of TFs in P. syringae. We also identified 35 vital TFs that participate in virulence pathways and 111 key TFs involved in metabolic pathways across the global transcriptional regulatory network in P. syringae. This analysis uncovered new functions of previously characterised TFs, such as MexT. In addition to the previously reported mexEF-oprN operon,(Sawada et al., 2018) this study identified fleQ (a flagellar regulator) and shcF (a type III chaperone protein) as the targets of MexT. In addition, we investigated the functional evolution and potential intra-species variability of TFs in P. syringae, demonstrating the functional diversity of TFs among P. syringae species during their long course of evolution. Based on the aforementioned results, we established the P. syringae transcriptional regulatory network (PSTRnet) database, which contains detailed binding peak information for all TFs in P. syringae (https://jiadhuang0417.shinyapps.io/PSTF_NET/). This database serves as a valuable platform for presenting, searching and downloading regulatory information of transcription in P. syringae.

Global transcriptional regulatory network in P. syringae.

The integrated transcriptional regulatory network reflects the interactions between all TFs classified into 39 families from different DNA binding domain types and target genes annotated from pathway annotations. The targets are shown in 12 pathways with various colors. TF nodes are gray as corresponding sizes representing TF number in the family. Edge colors are target-based.

We observed crosstalk between not only TFs but also various pathways. For example, the TF PSPPH4700 directly regulated fleQ and hrpR and indirectly regulated fleQ and hrpR through the PSPPH4700/PSPPH4324/PSPPH0755 cascade. Feedforward loop modules are usually coherent, meaning that the direct effect of downstream TFs has the same regulatory direction (positive or negative) as the indirect effect of upstream TFs(Shen-Orr et al., 2002). This observation enhanced our understanding of the influence of TFs in the network on their target genes based on the identified effects of other TFs. Furthermore, the TF PSPPH4700 was identified to bind to the promoters of genes involved in seven metabolic pathways, namely the TCA cycle, oxidative phosphorylation, methyl-accepting, amino acid biosynthesis, RNA polymerase, ABC transportation and DNA replication. Our results proved the possibility of crosstalk between different pathways in P. syringae.

Taken together, our study provides comprehensive insights into the DNA-binding characteristics and potential regulatory pathways of almost all annotated TFs in P. syringae. The global transcriptional regulatory network can not only contribute to the development of novel drugs to combat P. syringae infections but also advance research on the molecular mechanisms of TFs in other pathogens.

Materials and Methods

Bacterial strains, culture media, plasmids, and primers

The bacterial strains, plasmids and primers used in this study are listed in Supplementary Table 5. P. syringae 1448A, DC3000, B728a and C48 strains were grown at 28 °C in King’ B (KB) medium shaking at 220 rpm or KB agar plates(King, Ward, & Raney, 1954). E. coli BL21(DE3) or DH5α strains were grown at 37 °C in Luria-Bertani medium (LB) shaking at 220 rpm or on LB agar plates. Antibiotics used for P. syringae 1448A, DC3000 and B728a wide type (WT) strains and mutants were rifampicin at 50 μg/ml; P. syringae 1448A, DC3000 and B728a overexpression strains were rifampicin at 50 μg/ml and spectinomycin at 50 μg/ml, C48 overexpression strains were spectinomycin at 50 μg/ml; P. syringae 1448A strains with pK18mobsacB plasmids for mutant construction were rifampicin at 50 μg/ml and kanamycin at 50 μg/ml. Antibiotics used for E. coli with pET28a plasmids for protein purification, and pK18mobsacB plasmids were kanamycin at 50 μg/ml; E. coli with pHM1 plasmids for overexpression strain construction were spectinomycin at 50 μg/ml.

Construction of overexpression strains and mutants

Overexpression strains and mutants of P. syringae were constructed as previously described(Kvitko & Collmer, 2011; Shao et al., 2021). In brief, for overexpression strain, the open reading frame (Caillet, Baron, Boni, Caillet-Saguy, & Hajnsdorf) of each TF-coding gene was amplified by PCR from P. syringae genome and cloned into pHM1 plasmid. The ligated fragments were inserted into the digested pHM1 plasmids (HindIII) using ClonExpress MultiS One Step Cloning Kit (Vazyme). The recombinant plasmids were transformed into the P. syringae wild-type strain. The single colonies were confirmed by western blot. For mutants, the upstream (∼1500-bp) and downstream (∼1000-bp) fragments of TF ORF were amplified by PCR from the P. syringae genome and digested with XbaI respectively, and then ligated by T4 DNA ligase (Jumper et al.). The ligated fragments were inserted into the digested pK18mobsacB plasmids (XbaI) using ClonExpress MultiS One Step Cloning Kit (Vazyme). The recombinant plasmids were transformed into the P. syringae wild-type strain. The single colonies were selected to the sucrose plates, and further screened in two kinds of KB plates (with kanamycin/rifampin and with rifampin) concurrently. The single colonies losing kanamycin resistance were further verified by real-time quantitative PCR (RT-qPCR) to detect the mRNA level of corresponding TF genes.

ChIP-seq analyses

As previous description(Blasco et al., 2012), the overexpression strains of corresponding TFs and P. syringae WT with empty pHM1 plasmid was cultured in 30 mL KB medium to OD600=0.6. Bacterial cultures were cross-linked with 1% formaldehyde for 10 min at 28°C and then the reaction was stopped by the addition of 125 mM glycine for 5 min. The centrifugated bacteria were washed twice with Tris Buffer (20 mM Tris-HCl [pH 7.5] and 150 mM NaCl) and washed again with IP Buffer (50 mM HEPES–KOH [pH 7.5], 150 mM NaCl, 1 mM EDTA, 1% Triton X-100, 0.1% sodium deoxycholate, 0.1% SDS, and mini-protease inhibitor cocktail).

The centrifugated bacteria were preserved at 80℃ or continued for the next experiments. The bacteria were subjected with IP Buffer and then sonicated to pull down the DNA fragments (150-300-bp). The supernatant was the DNA-TF-HA tag complex and used as IP samples. IP experiments and control sample were incubated with agarose conjugated anti-HA antibodies (Sigma) in IP Buffer. The complex of DNA-TF-anti-HA agarose was applied to washing, crosslink reversal, and purification(Blasco et al., 2012). The 150-250-bp DNA fragments were selected for library construction. The libraries were sequenced using the HiSeq 2000 system (Illumina). Two biological replications have been performed for all ChIP-seq experiments. ChIP-seq reads were mapped to the P. syringae 1448A genome (NC_005773.3), DC3000 (NC_004578.1), B728a (NC_007005.1) and C48 (NZ_CP032631.1) using Bowtie2 (version 2.3.4.3)(Zhang et al., 2008). Subsequently, binding peaks (q < 0.01) were identified using MACS2 software (version 2.1.0). The enriched loci for each TF were annotated using the R package ChIPpeakAnno (version 3.18.2)(L. J. Zhu et al., 2010).

Electrophoretic mobility-shift assay

DNA probes were amplified from P. syringae 1448A genome by PCR using primers listed in Supplementary Table 5. The probes (20 ng) were mixed with various concentrations of proteins in 20 μL of gel shift buffer (10 mM Tris-HCl, pH 7.4, 50 mM KCl, 5 mM MgCl2, 10% glycerol). After incubation at room temperature for 20 min, the samples were analyzed by 6% native polyacrylamide gel electrophoresis (90 V for 90 min for sample separation). The gels were subjected to Gel Red dye (Tiangen Biotech) for 5 min and photographed by using a gel imaging system (Bio-Rad). The assay was repeated at least three times with similar results.

RT-qPCR

The RT-qPCR primers used are shown Supplementary Table 5 in the Supplemental Information. The cultured bacteria were grown to OD600=0.6 and the total RNA were purified with Bacteria Total RNA Isolation Kit (Sangon Biotech). The RNA concentrations were measured using a NanoDrop 2000 spectrophotometer (ThermoFisher). cDNA synthesis was performed using HiScript III RT SuperMix (Vazyme, China). RT-qPCR was performed with a SuperReal Premix Plus (SYBR Green) kit (Vazyme, China) according to the manufacturer’s instructions. The reactions used 100 nM primers and were run for 40 cycles at 95°C for 30 s and 95°C for 10 s, and at 60°C for 30 s. The fold change represents relative expression level of mRNA, which can be estimated by the values of 2-(ΔΔCt). The relative expression of target genes in WT was set to 1. All the reactions were conducted with three repeats.

Motility assay

The motility assay was performed based on our previous study(Shao et al., 2021). Swimming plates were KB agar plates containing 0.3% agar (MP Biomedicals, UK) and rifampicin at 25 μg/ml. Overnight bacterial cultures were inoculated on swimming plates as 2 μL aliquots and incubated at room temperature for 3-5 days. Finally, the diameter of motility trace represented the swimming motility of P. syringae strains. Photographs were taken by using the Bio-Rad imaging system. The assay was repeated at least three times with similar results.

Congo red assay

Congo red assay was performed as previous study(Shao et al., 2021). Congo red plates were KB agar plates containing 1.0% agar (MP Biomedicals, UK) and rifampicin at 25 μg/ml. Overnight bacterial cultures were inoculated on Congo red plates as 2 μL aliquots and incubated at 28℃. The colony staining was photographed after 5-7 days. The assay was repeated at least three times with similar results.

Biofilm formation assay

Biofilm formation assay was performed as previously describe(Shao et al., 2019). Overnight bacterial cultures were transferred to a sterile 10 mL borosilicate tube containing 2 ml KB medium with rifampicin (25 μg/ml) with the original concentration OD600 = 0.1. The cultures grew at room temperature for 36 hr without shaking. 0.1% crystal violet was used to stain the biofilm adhered to the tube for 20 min without shaking. Tubes were washed for more than three times with distilled deionized water gently and other components on the tube loosely was washed off. The tubes were dry and photographed. The remaining crystal violet was fully dissolved in 1 ml 95% ethanol with constantly shaking and measured its optical density at 590 nm (Biotek microplate reader). The assay was repeated at least three times with similar results.

Network and functional enrichment analysis

Network analyses were performed on Gephi 0.10. Functional annotations were retrieved from the Pseudomonas database and GO functional enrichment analyses and KEGG analysis were performed using DAVID v6.8.

Statistical analysis

Two-tailed Student’s t tests were performed using Microsoft Office Excel 2010. *P < 0.05, **P < 0.01, and ***P < 0.001 and results represent means ± SD. All experiments were repeated at least twice.

Data Availability

Sequencing data have been deposited and publicly available in Gene Expression Omnibus (GEO) under accession number GSE247395. Source data contain the numerical data used to generate the figures of EMSA. Hierarchical information and functional categories of TFs are available in Supplementary Table 1-3. Evolutionary details are provided in Supplementary Table 4. Primers and strains used in this paper are provided in Supplementary Table 5.

Funding

This study was supported by grants from the National Natural Science Foundation of China (31670127 and 31870116 to X.D.), General Research Funds of Hong Kong (21103018, 11101619, and 11102720 to X.D.). The funders had no role in study design, data collection, interpretation, or the decision to submit the work for publication.

eLife Assessment

This work advances our understanding of transcriptional regulation of virulence and metabolic pathways in plant pathogenic bacteria. Solid evidence for the claims is provided by computational analysis of the newly generated data on the genome-wide binding of 170 transcription factors to their target genes, together with experimental validation of the biological functions of some of these transcription factors. The findings and resources from this study will be valuable to researchers in the fields of systems biology, bacteriology, and plant-microbe interactions.

Reviewer #1 (Public Review):

In this work, the authors provide a comprehensive description of transcriptional regulation in Pseudomonas syringae by investigating the binding characteristics of various transcription factors. They uncover the hierarchical network structure of the transcriptome by identifying top-, middle-, and bottom-level transcription factors that govern the flow of information in the network. Additionally, they assess the functional variability and conservation of transcription factors across different strains of P. syringae by studying DNA-binding characteristics. These findings notably expand our current knowledge of the P. syringae transcriptome.

The findings associated with crosstalk between transcription factors and pathways, and the diversity of transcription factor functions across strains provide valuable insights into the transcriptional regulatory network of P. syringae. However, these results are at times underwhelming as their significance is unclear. This study would benefit from a discussion of the implications of transcription factor crosstalk on the functioning of the organism as a whole. Additionally, the implications of variability in transcription factor functions on the phenotype of the strains studied would further this analysis.

Overall, this manuscript serves as a key resource for researchers studying the transcriptional regulatory network of P. syringae.

Reviewer #2 (Public Review):

Summary:

The phytopathogenic bacterium Pseudomonas syringae is comprised of many pathovars with different host plant species and has been used as a model organism to study bacterial pathogenesis in plants. Transcriptional regulation is key to plant infection and adaptation to host environments by this bacterium. However, researchers have focused on a limited number of transcription factors (TFs) that regulate virulence-related pathways. Thus, a comprehensive, systems-level understanding of regulatory interactions between transcription factors in P. syringae has not been achieved.

This study by Sun et al performed ChIP-seq analysis of 170 out of 301 TFs in P. syringae pv. syringae 1448A and used this unique dataset to infer transcriptional regulatory networks in this bacterium. The network analyses revealed hierarchical interactions between TFs, various network motifs, and co-regulation of target genes by TF pairs, which collectively mediate information flow. As discussed, the structure and properties of the P. syringae transcriptional regulatory networks are somewhat different from those identified in humans, yeast, and E. coli, highlighting the significance of this study. Further, the authors made use of the P. syringae transcriptional regulatory networks to find TFs of unknown functions to be involved in virulence-related pathways. For some of these TFs, their target specificity and biological functions, such as motility and biofilm formation, were experimentally validated. Of particular interest is the finding that despite conservation of TFs between P. syringae pv. syringae 1448A, P. syringae pv. tomato DC3000, P. syringae pv. syringae B728a, and P. syringae pv. actinidiae C48, some of the conserved TFs show different repertoires of target genes in these four P. syringae strains.

Strengths:

This study presents a systems-level analysis of transcriptional regulatory networks in relation to P. syringae virulence and metabolism, and highlights differences in transcriptional regulatory landscapes of conserved TFs between different P. syringae strains, and develops a user-friendly database for mining the ChIP-seq data generated in this study. These findings and resources will be valuable to researchers in the fields of systems biology, bacteriology, and plant-microbe interactions.

Weaknesses:

No major weaknesses were found, but some of the results may need to be interpreted with caution. ChIP-seq was performed with bacterial strains overexpressing TFs. This may cause artificial binding of TFs to promoters which may not occur when TFs are expressed at physiological levels. Another caution is applied to the interpretation of the biological functions of TFs. The biological roles of the tested TFs are based on in vitro experiments. Thus, functional relevance of the tested TFs during plant infection and/or survival under natural environmental conditions remains to be demonstrated.

Reviewer #3 (Public Review):

Summary:

This study aims to understand gene regulation of the plant bacterial pathogen Pseudomonas syringae. Although the function of some TFs has been characterized in this strain, a global picture of the gene regulatory network remains elusive. The authors conducted a large-scale ChIP-seq analysis, covering 170 out of 301 TFs of this strain, and revealed gene regulatory hierarchy with functional validation of some previously uncharacterized TFs.

Strengths:

- This study provides one of the largest ChIP-seq datasets for a single bacterial strain, covering more than half of its TFs. This impressive resource enabled comprehensive systems-level analysis of the TF hierarchy.

- This study identified novel gene regulation and function with validations through biochemical and genetic experiments.

- The authors attempted on broad analyses including comparisons between different bacterial strains, providing further insights into the diversity and conservation of gene regulatory mechanisms.

Weaknesses:

(1) Some conclusions are not backed by quantitative or statistical analyses, and they are sometimes overinterpreted.

(2) Some figures and analyses are not well explained, and I was not able to understand them.

(3) The Method section lacks depth, especially in data analyses. It is strongly recommended that the authors share their analysis codes so that others can reproduce the analyses.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

In this work, the authors provide a comprehensive description of transcriptional regulation in Pseudomonas syringae by investigating the binding characteristics of various transcription factors. They uncover the hierarchical network structure of the transcriptome by identifying top-, middle-, and bottom-level transcription factors that govern the flow of information in the network. Additionally, they assess the functional variability and conservation of transcription factors across different strains of P. syringae by studying DNA-binding characteristics. These findings notably expand our current knowledge of the P. syringae transcriptome.

The findings associated with crosstalk between transcription factors and pathways, and the diversity of transcription factor functions across strains provide valuable insights into the transcriptional regulatory network of P. syringae. However, these results are at times underwhelming as their significance is unclear. This study would benefit from a discussion of the implications of transcription factor crosstalk on the functioning of the organism as a whole. Additionally, the implications of variability in transcription factor functions on the phenotype of the strains studied would further this analysis.

Overall, this manuscript serves as a key resource for researchers studying the transcriptional regulatory network of P. syringae.

Thank you for your positive comments.

Reviewer #2 (Public Review):

Summary:

The phytopathogenic bacterium Pseudomonas syringae is comprised of many pathovars with different host plant species and has been used as a model organism to study bacterial pathogenesis in plants. Transcriptional regulation is key to plant infection and adaptation to host environments by this bacterium. However, researchers have focused on a limited number of transcription factors (TFs) that regulate virulence-related pathways. Thus, a comprehensive, systems-level understanding of regulatory interactions between transcription factors in P. syringae has not been achieved.

This study by Sun et al performed ChIP-seq analysis of 170 out of 301 TFs in P. syringae pv. syringae 1448A and used this unique dataset to infer transcriptional regulatory networks in this bacterium. The network analyses revealed hierarchical interactions between TFs, various network motifs, and co-regulation of target genes by TF pairs, which collectively mediate information flow. As discussed, the structure and properties of the P. syringae transcriptional regulatory networks are somewhat different from those identified in humans, yeast, and E. coli, highlighting the significance of this study. Further, the authors made use of the P. syringae transcriptional regulatory networks to find TFs of unknown functions to be involved in virulence-related pathways. For some of these TFs, their target specificity and biological functions, such as motility and biofilm formation, were experimentally validated. Of particular interest is the finding that despite conservation of TFs between P. syringae pv. syringae 1448A, P. syringae pv. tomato DC3000, P. syringae pv. syringae B728a, and P. syringae pv. actinidiae C48, some of the conserved TFs show different repertoires of target genes in these four P. syringae strains.

Thank you for your positive comments.

Strengths:

This study presents a systems-level analysis of transcriptional regulatory networks in relation to P. syringae virulence and metabolism, and highlights differences in transcriptional regulatory landscapes of conserved TFs between different P. syringae strains, and develops a user-friendly database for mining the ChIP-seq data generated in this study. These findings and resources will be valuable to researchers in the fields of systems biology, bacteriology, and plant-microbe interactions.

Thank you for your positive comments.

Weaknesses:

No major weaknesses were found, but some of the results may need to be interpreted with caution. ChIP-seq was performed with bacterial strains overexpressing TFs. This may cause artificial binding of TFs to promoters which may not occur when TFs are expressed at physiological levels. Another caution is applied to the interpretation of the biological functions of TFs. The biological roles of the tested TFs are based on in vitro experiments. Thus, functional relevance of the tested TFs during plant infection and/or survival under natural environmental conditions remains to be demonstrated.

Thank you for your comments, and we agree with the reviewer. To eliminate the artificial binding of TFs, we performed EMSA to verify the analyzed targets. Our EMSA results confirmed the analyzed binding peaks.

For the verification experiments of the biological functions of TFs, we also performed in vivo motility assay and biofilm production assay (Figures 3b-d). To further detect the biological functions of TFs, we performed plant infection assay of TF PSPPH2193 under natural environmental condition (bean leaves). As shown in Figures S6c and g, both the motility and the virulence of P. syringae in ∆PSPPH2193 strain was significantly reduced compared with WT strain. These results showed that TF PSPPH2193 positively regulated the pathogenicity of P. syringae via modulating the bacterial motility.

Reviewer #3 (Public Review):

Summary:

This study aims to understand gene regulation of the plant bacterial pathogen Pseudomonas syringae. Although the function of some TFs has been characterized in this strain, a global picture of the gene regulatory network remains elusive. The authors conducted a large-scale ChIP-seq analysis, covering 170 out of 301 TFs of this strain, and revealed gene regulatory hierarchy with functional validation of some previously uncharacterized TFs.

Thank you for your positive comments.

Strengths:

- This study provides one of the largest ChIP-seq datasets for a single bacterial strain, covering more than half of its TFs. This impressive resource enabled comprehensive systems-level analysis of the TF hierarchy.

- This study identified novel gene regulation and function with validations through biochemical and genetic experiments.

- The authors attempted on broad analyses including comparisons between different bacterial strains, providing further insights into the diversity and conservation of gene regulatory mechanisms.

Thank you for your positive comments.

Weaknesses:

(1) Some conclusions are not backed by quantitative or statistical analyses, and they are sometimes overinterpreted.

Thank you for your comments. We used hypergeometric test in this analysis. Although only one gene was enriched in some pathways, the adjusted p-value was less than 0.05. We added the details in the revised manuscript.

(2) Some figures and analyses are not well explained, and I was not able to understand them.

Thank you for your comments, and we are sorry for the confusion. We defined ‘indirect interaction’ as ‘co-association’ and ‘cooperativity’ as ‘if the common target of two TFs is from a TF’. We added the definition of "indirect interaction" and "cooperativity" in the revised legend.

For Figure S3a, the low co-association scores and large peak numbers of these top-level TFs indicated that top-level TFs preferred to solely regulate target genes, but not to co-regulate with other top-level TFs. PSPPH4700 was an example to show that top-level TFs with low co-association scores and large peak numbers tend to solely regulate target genes, but not to co-regulate with other top-level TFs. We revised the sentence to ‘For example, the top-level TF PSPPH4700 yielded over 1,700 peaks but cooperated with only 24 top-level TFs with low co-association scores about 0.05 (Supplementary Table 2b).’.

We analyzed high co-association scores of 125 TFs in three levels and further determined the co-association patterns. To identify the tendency of co-association of all these 125 TFs, the co-association patterns were classified into 4 clusters. Bottom-level TFs tend to co-regulate target genes with other TFs. We revised the sentence in the revised manuscript.

For Figure 2b, in C1, C2 and C4, many bottom-level TFs performed co-association pattern with other TFs, especially bottom TFs (showed in C4). To explore the regulatory pattern in C3, the peak locations in target genes of MexT were analyzed with those of TFs in C3. Seven top-level TFs (PSPPH1435, PSPPH1758, PSPPH2193, PSPPH2454, PSPPH4638, PSPPH4998 and PSPPH3411), three middle-level TFs (PSPPH1100, PSPPH5132 and PSPPH5144) and four bottom-level TFs (PSPPH0700, PSPPH2300, PSPPH2444 and PSPPH2580) were compared with MexT. MexT showed higher co-association scores (more than 60 scores) with more top-level-TFs. Therefore, we demonstrated that MexT performed closer co-association relationships with top-level TFs. We added the statement in the revised manuscript.

For Figure 1a, the hierarchical network showed different number of TFs in three levels (54 top-level TFs, 62 middle-level TFs and 147 bottom-level TFs), which indicated that more than half of TFs (bottom-level TFs) tend to be regulated by other TFs and then directly bound to target genes. This finding showed a downward regulatory direction of transcription regulation in P. syringae. We revised the statement in the revised manuscript.

(3) The Method section lacks depth, especially in data analyses. It is strongly recommended that the authors share their analysis codes so that others can reproduce the analyses.

Thank you for your comments, and we defined the intergenic region before each TF sequence as the promoter region. As pHM1 plasmid carries its own constitutive promoter (lacZ promoter), we amplified the TF-coding sequence and cloned into site following the promoter. The TF protein expression was activated by the promoter of plasmid. Psph 1448A was used for our main ChIP-seq. We added the details in the revised manuscript.

For Figure S3, we performed GO analysis on genes that were co-bound by TF pairs. We added the details in the revised manuscript.

We shared our analysis codes on the website (https://github.com/dengxinb2315/PS-PATRnet-code) in the Data Availability.

Recommendations for the authors

Reviewer #1 (Recommendations For The Authors):

(1) The specific strain of Pseudomonas syringae used in the study outside of the evolutionary analysis should be specified in the abstract and main text.

Thank you for your suggestion. We revised the statements in abstract and main text to specific strains.

(2) The language used throughout the manuscript should be revised for clarity, conciseness, and readability.

Thank you for your suggestion. We have revised the language used throughput the manuscript by a scientific editor who is a native speaker of English.

(2) Line 688: Replace "80C" with "-80C".

Thank you for your correction. We revised ‘80℃’ to ‘-80℃’. Please see Line 713.

(3) Line 172 - 173: The abbreviations TT, MM, BB, TM, TB, and MB need to be expanded in the main text before their use.

Thank you for your suggestion. We added the abbreviations TT, MM, BB, TM, TB, and MB in the manuscript. Please see Lines 172-174.

Reviewer #2 (Recommendations For The Authors):

Major points

(1) The name of the P. syringae strains used in each experiment/analysis should be explicitly stated (most experiments were carried out with P. syringae strain 1448A). This should also be applied to the introduction where many papers on P. syringae are cited without clear indication of strain names. I think this amendment is essential because target genes and thus biological functions of TFs could be different between P. syringae strains, as shown in the present study.

Thank you for your suggestion. We revised the P. syringae strains in the citations throughout the manuscript.

(2) How many TFs were analyzed throughout the study? Most sentences including line 22 in the abstract say 170, but I also found some say 270 (for example, line 106 and line 149). The legend of Figure 1 says 262. More detailed information is required regarding the datasets used for each analysis.

Thank you for your suggestion. The number of TFs analyzed by ChIP-seq in this research is 170, the number of TFs analyzed by HT-SELEX in our previous research is 100. Hierarchical analysis integrated data from ChIP-seq and HT-SELEX which included 270 TFs. As 8 TFs did not show hierarchical characteristic, the legend of Figure 1 said 262 TFs. We added the data source in the revised manuscript. Please see Lines 104, 147, 160 and 1082.

(3) Figure 1b: Please define "indirect interaction" and "cooperativity" in the legend as well as in the text. I only found the definition of "direct interaction".

Sorry for the missing information. We defined ‘indirect interaction’ and ‘cooperativity’ as ‘co-association’ and ‘if the common target of two TFs is from a TF’, respectively. We added the definition of "indirect interaction" and "cooperativity" in the revised legend. Please see Lines 174-176, 1084-1086.

(4) I found it very interesting that conserved TFs show different repertoires of target genes in different P. syringae strains. This suggests the rewiring of transcriptional regulatory networks in P. syringae strains, but the underlying mechanism is not explored in the current manuscript. It can be easily tested whether these conserved TFs bind to similar or different motifs by motif enrichment analysis. If they bind to similar motifs, it is possible that the promoter sequences of their target genes have diversified. Addressing or at least discussing these points would provide molecular insights into the diversification of the transcriptional regulatory networks in P. syringae. Similarly, functional enrichment analysis of target genes can be used to test whether the conserved TFs regulate different biological processes.

Thank you for your suggestion. We added the motif analysis and functional enrichment analysis of target genes of TFs (PSPPH3122 and PSPPH4127) in different P. syringae strains. We found two different motifs (AGACN4GATCAA and CGGACGN3GATCA) in 1448A and DC3000 strains, respectively. We also performed the GO analysis and found the specific functions of PSPPH3122 in Psph 1448A compared with Pst DC3000 and Pss B728a strains, including recombinase activity and DNA recombination. For PSPPH4127, we found four different motifs in four P. syringae strains. GO analysis showed its relationship with recombinase activity in Psph 1448A strain, and RNA binding, structural constituent of ribosome, translation and ribosome in Pss B728a strain. These results indicated the highly functional diversity of TFs in P. syringae. We added these points in the Results part, and Figure S9-S10 in the revised manuscript. Please see Lines 497-509.

(5) Related to point 4, it would be quite useful if a list of orthologous genes of 1448A TFs in the other tested P. syringae strains were provided. Such information may also enhance the utility of the database developed in this study.

Thank you for your suggestion. We added the list of orthologous genes of 301 Psph 1448A TFs in the other tested P. syringae strains in the Supplementary Table 5. Please see Lines 467 and Supplementary Table 5.

(6) Lines 243-246: It is unclear how these functional enrichment analyses were performed. Did you use target genes regulated by individual TFs or those coregulated by pairs of TFs? Please add more information for the sake of readers.

Thank you for your suggestion. We performed the functional enrichment analyses by hypergeometric test (BH-adjusted p < 0.05) via using target genes regulated by individual TFs. We added the details in the Results part. Please see Lines 248-252, 270, 1194-1195, 1199-1200 and 1205-1206.

Minor points

(1) Lines 167-168: I may not understand correctly, but you might want to say "downward-pointing edges" instead of "upward-pointing edges".

Thank you for correction. We revised the ‘upward-pointing edges’ to ‘downward-pointing edges’. Please see Line 166.

(2) Line 174: "physical interactions" should be amended to "direct interactions".

Thank you for correction. We revised the ‘physical interactions’ to ‘direct interactions’. Please see Line 177.

(3) Line 224: Could you please explain why bacterial growth in plant tissues is considered an example of "multi-stability"?

Thank you for your suggestion. We are sorry for the incorrect statement. We showed ‘plant intercellular spaces’ as ‘multi-stability’. We revised the sentence to ‘These auto-regulators are important and always act as repressors in scenarios of multi-stability, such as plant intercellular spaces’. Please see Lines 224-226.

(4) Line 254-257: Here, the definition of "tether binding" is introduced, but it is not very clear to me. In my understanding, tethered binding is an indirect binding of a TF to a target gene through protein-protein interaction with other TF that directly binds to the promoter of the target gene.

Thank you for your suggestion, and we agree with you. We referred to the paper published in 2012 (Wang et al., 2012) and revised the statement of ‘tether binding’ to ‘This finding suggested that these TFs indirectly regulated target genes through protein-protein interaction with other TFs that directly binds to the promoters of target genes, a phenomenon defined as tethered binding’. Please see Lines 259-262.

(5) Lines 341-343: Figure 3b shows qRT-PCR of hopAE1, not hrpR.

Thank you for your correction. We revised ‘hrpR’ to ‘hopAE1’. Please see Line 349.

(6) Lines 500 and Figure 6b: It is hard to see edges from module 12 to others. So, it would be better to provide numeric information (number of TFs and target genes) in the text.

Thank you for your suggestion. Module 12 includes 22 TFs and 318 target genes. We added the statement of numeric information about Module 12 in the revised manuscript. Please see Lines 536-537.

(7) Line 519: Figure S4b is not the EMSA data for PSPPH3798. Should it be Figure S4e?

Thank you for your correction. We revised to ‘Figure S4e’. Please see Line 545.

(8) Line 522: Figure S6b is not relevant to the statement here.

Thank you for your correction. We deleted the ‘Figure S6b’ here. Please see Line 547.

(9) Line 593: prokaryotic transcriptional regulatory networks -> eukaryotic transcriptional regulatory networks?

Thank you for your correction. We revised ‘prokaryotic transcriptional regulatory networks’ to ‘eukaryotic transcriptional regulatory networks’. Please see Line 618.

(10) Figure S3 requires images of higher resolution. Especially, values for the color codes are not readable or very hard to see.

Thank you for your suggestion. To make the images clearer, we enlarged the images, change the color codes, and divided it into three figures. Please see the revised Figures S3-S5 and corresponding Figure legends at Lines 1191-1206.

**Reviewer #3 (Recommendations For The Authors):

**(1) Some conclusions are not backed by quantitative or statistical analyses, and they are sometimes overinterpreted.

L221: "Taken together, the simplest and most effective submodule M1 and the coregulatory submodule M13 played crucial roles in the transcriptional regulation of TFs in P. syringae."

The authors did not provide any evidence supporting the functional importance of any of these submodules. M13 is most enriched within the locked loop, but its size is much smaller than simple loops. What evidence supports the importance of this particular submodule?

Thank you for your suggestion. In eukaryote (Saccharomyces cerevisiae) and prokaryote (Escherichia coli) which have the best characterized transcriptional regulation networks, the feed-forward loop (called M13 in this article) appear numerous times in the networks and perform different biological functions. M1 appeared most frequently by an order of magnitude than other modules. We revised the sentence to ‘Taken together, the most numerous but simplest submodule M1 played a crucial role in the transcriptional regulation of TFs in P. syringae.’ Please see Lines 222-224.

L223: "...we found 92 auto-regulators...These auto-regulators are important and always act as repressors in scenarios of multi-stability, such as in plant intercellular spaces where bacteria grow (Figure 1d)(Alon, 2007). These regulators are regarded as bistable switches that further influence the expression of downstream genes."

Are these claims supported by any evidence?

Thank you for your suggestion. We referred to the following articles:

(1) Alon. Nature Reviews Genetics. 2007(Alon, 2007).

That transcription factors repress the transcription of their target genes was considered as negative regulation. These negative autoregulators account for half of the repressors in E. coli and occur in many eukaryotes. The repressors controlled the concentration of the target production through suppressing its expression, which accelerated back to the steady state of cells.

(2) Becskei. et al. Nature. 2000; Rosenfeld et al. Journal of Molecular Biology. 2002 (Becskei & Serrano, 2000; Rosenfeld, Elowitz, & Alon, 2002).

Fluorescent assay confirmed that the negative autoregulatory module (negative autoregulator TetR) spent less time to the log phase than unregulated group, which reduced cell-to-cell fluctuations in the steady-state level of the transcription factor. Some negative autoregulators were showed here, such as LexA, CysB and SrlA-D.

In our research, we also identified many autoregulators including CysB and LexA2 (annotated as LexA repressor). We revised the sentence to ‘In addition, we found 92 auto-regulators in our hierarchy network. These auto-regulators are important and always act as repressors in scenarios of multi-stability, such as plant intercellular spaces (Figure 1d) (Alon, 2007). For example, LexA and CysB as negative autoregulators were indicated to reduce cell-to-cell fluctuations in the steady-state level of the transcription factor (Becskei & Serrano, 2000; Rosenfeld et al. 2002).’. Please see Lines 224-229.

L265: "This finding indicated that the bottom-level TFs, which were more easily regulated, tended to cooperate with downstream genes and other intra-level TFs."

Could the authors provide more explanation to reach this conclusion from the data? Analyzing the number of highly co-accessing TFs does not sufficiently support this conclusion. The clustering of TFs (C1-C4) is incomplete, and each TF level (Top/Middle/Bottom) contains different numbers of TFs. Since the authors calculated all-by-all co-association scores for these 125 TFs, they can group these scores into 6 possible combinations (TT, TM, TB, MM, MB, BB) and show the distribution of co-association scores.

Thank you for your suggestion. We indicated that the bottom-level TFs preferred to regulate the target genes through the cooperation with other TFs. To further support the claim, we analyzed the proportion of the bottom TF interaction in all the TF pairs interactions and direct interaction based on results in Figure 1B. The interactions of bottom TFs were 43% and 49%, respectively. However, the interactions of top TFs and middle TFs were only 20% and 28%, respectively. We revised the statement ‘Based on the analysis in Figure 1B, we found that the proportions of bottom-level TF interaction in all the TF pair interactions and direct interaction were 43% and 49%. These results indicated that the bottom-level TFs tended to regulate downstream genes through cooperating with other level TFs.’ in the revised manuscript. Please see Lines 269-272.

As not every TF performed co-association with other TFs, we only collected 125 TFs with co-association scores. For the numbers of TF in each level, we divided TFs into three levels according to hierarchy height. Hierarchy height from -1 to -0.3 represented bottom level; hierarchy height from -0.3 to 0.3 represented middle level ; hierarchy height from 0.3 to 1 represents top level. Each level was equally divided by height scores. We suggested that different numbers of TFs in three levels indicated the characteristic of transcriptional regulation in P. syringae.

Thank you for your suggestion. As the co-association patterns were determined by co-association scores of the same TFs, we first grouped the co-association scores into 3 possible TF pairs (TT, MM, and BB, in Figures S3a, S4a and S5a). Our results indicated that higher co-association scores preferred to occur in bottom-level TFs. We revised the statement in the revised manuscript. Please see Lines 244-252.

(2) Some figures and analyses are not well explained, and I was not able to understand them.

Figure 1b: The terms "direct," "indirect," and "cooperativity" require further clarification as their definitions in the text (L169-183) are unclear to me. This ambiguity hampers the evaluation of the authors' discussion regarding TF-TF interactions (L561-584), an important theme of this study. The figure includes concepts discussed in later sections (e.g., cooperativity), making it difficult to understand. A diagram explaining these concepts would be highly helpful for readers to understand.

Sorry for the missing information. We defined ‘indirect interaction’ as ‘co-association’, ‘cooperativity’ as ‘if the common target of two TFs is from a TF’. We added the definition of "indirect interaction" and "cooperativity" in the revised manuscript and legend. Please see Lines 174-176 and 1085-1087.

L253: "Notably, we found that TFs at the top level, without cooperating TFs, exhibited a large number of binding peaks (Figure S3a)."

I could not understand this sentence. Did the authors mean that top-level TFs with a large number of peaks showed a low level of co-association? If so, does this data suggest that these TFs do not tend to cooperate with other TFs? I was confused by the discussion in L253-L261.

Thank you for your comment, and we agree with you. The low co-association scores and large peak numbers of these top-level TFs indicated that top-level TFs preferred to solely regulate target genes, but not to co-regulate with other top-level TFs.

Thank you for your comment. From L253-256, PSPPH4700 was an example to show that top-level TFs with low co-association scores and large peak numbers tend to solely regulate target genes, but not to co-regulate with other top-level TFs. We revised the sentence to ‘For example, the top-level TF PSPPH4700 yielded over 1,700 peaks, but cooperated with only 24 top-level TFs with low co-association scores about 0.05 (Supplementary Table 2b).’.

From L257-261, we analyzed high co-association scores of 125 TFs in three levels and further determined the co-association patterns. To identify the tendency of co-association of all these 125 TFs, the co-association patterns were classified into 4 clusters. Bottom-level TFs tend to co-regulate target genes with other TFs. We revised the sentence. Please see Lines 262-264, 265-266 and 269-272.

L287: "The analysis of the peak locations of MexT demonstrated that MexT showed closer co-association relationships with top-level TFs (Figure 2b)."

I could reach this conclusion by seeing Figure 2b. Additional explanation and/or data visualization would be appreciated.

Thank you for your suggestion. In C1, C2 and C4, many bottom-level TFs performed co-association pattern with other TFs, especially bottom TFs (showed in C4). To explore the regulatory pattern in C3, the peak locations in target genes of MexT were analyzed with those of TFs in C3. Seven top-level TFs (PSPPH1435, PSPPH1758, PSPPH2193, PSPPH2454, PSPPH4638, PSPPH4998 and PSPPH3411), three middle-level TFs (PSPPH1100, PSPPH5132 and PSPPH5144) and four bottom-level TFs (PSPPH0700, PSPPH2300, PSPPH2444 and PSPPH2580) were compared with MexT. MexT showed higher co-association scores (more than 60 scores) with more top-level-TFs. Therefore, we demonstrated that MexT performed closer co-association relationships with top-level TFs. We added the statement in the revised manuscript. Please see Lines 291-296.

Figure 6cd: What kind of enrichment analysis did the authors perform? Was any statistical test used? The figure only shows the number of genes, and sometimes the number is only 1 for a functional category. Can it be considered as significant enrichment?

Thank you for your comment. We used hypergeometric test in this analysis. Although only one gene was enriched in some pathways, the adjusted p-value was less than 0.05. We added the details in the revised manuscript. Please see Lines 533-534.

L169: "The hierarchical network revealed a downward information flow, suggesting the prioritization of collaboration between different hierarchy levels."

Can the authors please explain the logic behind this statement more in detail?

Thank you for your comment. The hierarchical network showed different number of TFs in three levels (54 top-level TFs, 62 middle-level TFs and 147 bottom-level TFs), which indicated that more than half of TFs (bottom-level TFs) tend to be regulated by other TFs and then directly bound to target genes. This finding showed a downward regulatory direction of transcription regulation in P. syringae. We revised the statement in the revised manuscript. Please see Lines 167-170.

(3) The Method section lacks depth, especially on data analyses.

How did the authors define promoter regions of each gene? How were operons treated in their analyses? Was P. syringae 1448A used for their main ChIP-seq?

Thank you for your comment. We defined the intergenic region before each TF sequence as the promoter region.

As pHM1 plasmid carries its own constitutive promoter (lacZ promoter), we amplified the TF-coding sequence and cloned into the site following the promoter. The TF protein expression was activated by the promoter of plasmid.

P. syringae 1448A was used for our main ChIP-seq. We added the details in the revised manuscript. Please see Lines 705 and 727-730.

Figure S3: I am not sure how the GO analyses were done. For example, in the case of the top-level TF PSPPH4700, did the authors perform GO analysis on genes that are co-bound by PSPPH4700 and any other top-level TFs?

Thank you for your comment and we agree with you. We performed GO analysis on genes that were co-bound by TF pairs in the same level. We added the details in the revised manuscript. Please see Lines 248-252.

The analysis presented in Figure 6a needs more explanation of the methodology employed by the authors.

Thank you for your comment. We added more details for the analysis in Figure 6a. Please see Lines 514-522.

It is strongly recommended that the authors share their analysis codes so that others can reproduce the analyses.

Thank you for your comment. We shared our analysis codes on the website (https://github.com/dengxinb2315/PS-PATRnet-code) in the Data Availability. Please see Lines 800-801.

(4) Other:

Figure 3: I suggest putting additional panel labels to facilitate the interpretation of the figure.

Thank you for your suggestion. We added detailed labels in the revised Figures 3 and 4. Please see in the revised Figures 3 and 4.

I spotted several potential errors:

L106: 170 TFs?

Thank you for your comment, and we are sorry for the missing details. For the hierarchical network, we integrated the DNA-binding data of 170 TFs in this study and 100 TFs in our previous SELEX research. We added the details in the revised manuscript. Please see Lines 104, 147 and 159-160.

L592: P. syringae not E. coli?

Thank you for your comment. Here we discussed the hierarchical characteristics in E. coli. We revised the statement in the revised manuscript. Please see Line 618.

L593: eukaryotic not prokaryotic?

Thank you for your correction. Here we discussed the feedforward loops in our study. We revised the statement in the revised manuscript. Please see Line 618.

References

Alon, U. (2007). Network motifs: theory and experimental approaches. Nature Reviews Genetics, 8(6), 450-461.

Becskei, A., & Serrano, L. (2000). Engineering stability in gene networks by autoregulation. Nature, 405(6786), 590-593.

Rosenfeld, N., Elowitz, M. B., & Alon, U. (2002). Negative autoregulation speeds the response times of transcription networks. Journal of molecular biology, 323(5), 785-793.

Wang, J., Zhuang, J., Iyer, S., Lin, X., Whitfield, T. W., Greven, M. C., . . . Cheng, Y. (2012). Sequence features and chromatin structure around the genomic regions bound by 119 human transcription factors. Genome research, 22(9), 1798-1812.

Version 2

Abstract

Abstract

The model Gram-negative plant pathogen Pseudomonas syringae utilises hundreds of transcription factors (TFs) to regulate its functional processes, including virulence and metabolic pathways that control its ability to infect host plants. Although the molecular mechanisms of regulators have been studied for decades, a comprehensive understanding of genome-wide TFs in Psph 1448A remains limited. Here, we investigated the binding characteristics of 170 of 301 annotated TFs through ChIP-seq. Fifty-four TFs, 62 TFs and 147 TFs were identified in top-level, middle-level and bottom-level, reflecting multiple higher-order network structures and direction of information-flow. More than forty thousand TF-pairs were classified into 13 three-node submodules which revealed the regulatory diversity of TFs in Psph 1448A regulatory network. We found that bottom-level TFs performed high co-associated scores to their target genes. Functional categories of TFs at three levels encompassed various regulatory pathways. Three and 25 master TFs were identified to involve in virulence and metabolic regulation, respectively. Evolutionary analysis and topological modularity network revealed functional variability and various conservation of TFs in P. syringae (Psph 1448A, Pst DC3000, Pss B728a and Psa C48). Overall, our findings demonstrated the global transcriptional regulatory network of genome-wide TFs in Psph 1448A. This knowledge can advance the development of effective treatment and prevention strategies for related infectious diseases.

Introduction

Transcription is a pivotal process in cellular life events. Transcription factors (TFs) play a crucial role in this process by acting as key regulators that coordinate various biological activities(Lee & Young, 2013; Papavassiliou & Papavassiliou, 2016). TFs control the recruitment of RNA polymerase by identifying and binding to the promoters of downstream genes, thereby either activating or repressing the expression of target genes(Lambert et al., 2018; Wade, 2015). Numerous studies have focused on regulatory events in eukaryotic species, such as humans(Jolma et al., 2013), mice(Badis et al., 2009), and Saccharomyces cerevisiae(C. Zhu et al., 2009), and prokaryotic species, such as Escherichia coli(Shen-Orr, Milo, Mangan, & Alon, 2002). However, limited comprehensive TF binding datasets for microbial pathogens are available.

Pseudomonas syringae, an important Gram-negative phytopathogen and a model pathogenic bacterium, infects many plants, including economically valuable crops, resulting in substantial annual economic losses globally(Hirano & Upper, 2000). Upon entering host cells, P. syringae employs several strategies, such as changing its motility type and secreting phytotoxins, to overcome the plant’s immune defences and establish colonies(Bender, Alarcón-Chaidez, & Gross, 1999; Ichinose, Taguchi, & Mukaihara, 2013; Taguchi & Ichinose, 2011). P. syringae causes severe disease by secreting various effector proteins through the needle-like type III secretion system (T3SS); this process is regulated by a cluster of TFs (Cunnac, Lindeberg, & Collmer, 2009; Hendrickson, Guevera, & Ausubel, 2000; Huang, Yao, Sun, Ji, & Deng, 2022; Jingru Wang et al., 2018). The alternative sigma factor RpoN activates the transcription of another alternative sigma factor, HrpL, which, in turn, binds to the pathogenicity (hrp) box in the promoter region of T3SS genes, regulating most of these T3SS genes (Alfano & Collmer, 1997; Lan, Deng, Zhou, & Tang, 2006; Yingxian Xiao & Hutcheson, 1994). HrpS is one of the most important TFs that regulate numerous biological processes (Jingru Wang et al., 2018). Its heterodimeric complex, HrpRS, is modulated by at least six two-component systems (TCSs): RhpRS (Deng et al., 2014), CvsRS (Fishman, Zhang, Bronstein, Stodghill, & Filiatrault, 2018), GacAS (Chatterjee et al., 2003), AauRS (Yan, Rogan, Pang, Davis, & Anderson, 2020), CbrAB2 and EnvZ-OmpR (Shao et al., 2021). In particular, RhpRS serves as a master regulator of T3SS in Psph 1448A. RhpRS senses plant-derived signals, such as polyphenols, through the histidine kinase Pro40 within RhpS and controls the expression of T3SS genes in response to environmental stress(Deng et al., 2014; Yanmei Xiao et al., 2007; Xie et al., 2021). Within the sensor region, the cognate response regulator RhpR undergoes modulation in its phosphorylation state by RhpS, thereby regulating a group of T3SS genes(Deng et al., 2010). Phosphorylated RhpR directly binds to the hrpRS promoter, suppressing the hrpRS operon and the subsequent hrpRS-hrpL-hrp cascade (Deng et al., 2010; Deng et al., 2014; Deng, Xiao, Lan, Zhou, & Tang, 2009; Shao, Xie, Zhang, & Deng, 2019; Yanmei Xiao et al., 2007).

Recently, through a combined analysis of RNA sequencing (RNA-seq) and chromatin immunoprecipitation sequencing (ChIP-seq), we identified seven additional TCSs (ErcS, Dcsbis, PhoBR, CzcSR, AlgB/KinB, MerS and CopRS) that regulate the virulence of Psph 1448A (Xie et al., 2022). In addition, we developed an intricate PSTCSome (Psph 1448A TCS regulome) network containing numerous functional genes that respond to changing environmental conditions(Xie et al., 2022). Furthermore, we examined the overall crosstalk between 16 virulence-related regulators under different growth conditions, such as King’s B and minimal media. By analysing differentially expressed genes and binding peaks, we constructed a Psph 1448A regulatory network (PSRnet), revealing the involvement of hundreds of functional genes in virulence pathways(Shao et al., 2021). We also elucidated the molecular mechanisms and functions of TFs binding within coding sequences (CDS) and found that CDS-binding TFs interact with cryptic promoters in coding regions, thereby regulating the expression of subgenus and antisense RNAs(Hua et al., 2022). We propose a luminescence reporter system designed to quantitatively measure the translational elongation rates (ERs) of T3SS-related proteins. Our findings demonstrate the key roles of transfer RNAs (tRNAs) and elongation factors in modulating translational Ers and facilitating T3SS protein synthesis(Sun et al., 2022).

Although many key virulence regulators in Psph 1448A have been studied, the global regulatory mechanism and interactions of all 301 annotated TFs across various biological processes remain unclear. To comprehensively explore the DNA-binding features and map the transcriptional regulatory network of all TFs in Psph 1448A, we constructed 170 TF-overexpressing strains and used ChIP-seq, a highly effective and important technology for analysing protein–DNA interactions(Mathelier, Shi, & Wasserman, 2015). This analysis not only provided insights into the interactions between TFs and their target genes but also revealed the hierarchy (top, middle and bottom) and co-association scores of all these TFs. We found that more than half of 270 TFs (100 TFs from HT-SELEX and 170 TFs this study) in downstream position tended to be regulated by top TFs and bound to the target genes with high co-associated scores. Different TF-pairs were classified into 13 basic three-node submodules, including ringent loops and locked loops. In addition, we mapped the hierarchical binding network of TFs and identified three virulence-related master TFs and 23 metabolic master TFs. Furthermore, we employed ChIP-seq to determine the binding sites of 5 TFs in 4 P. syringae lineages (Psph 1448A, Pst DC3000, Pss B728a and Psa C48), revealing the diversity of TF binding events and the varying functions of TFs among different P. syringae strains. Topological modularity classification of the network, including TFs and target genes, revealed the diverse biological functions of TFs in Psph 1448A. This study provides a global and convenient platform for understanding the transcriptional regulatory characteristics and biological functions of TFs in P. syringae. In addition, this study provides valuable insights that can inform the development of effective therapies for not only P. syringae but also other associated infectious diseases.

Results

ChIP-seq analysis revealed the binding specificities of 170 previously uncharacterised TFs in Psph 1448A

Based on the current annotations available on ‘Pseudomonas Genome DB’ (https://www.pseudomonas.com/)(Winsor et al., 2016), we initially determined the locations of all 301 annotated TFs in the Psph 1448A genome (Figure S1a). To elucidate the binding preferences and functional characteristics of TFs of Psph 1448A, we performed ChIP-seq for the 170 TFs, including three (1.8%) predicted transcriptional regulators, 132 (77.6%) annotated transcriptional regulators and 35 (20.6%) functional proteins with DNA-binding annotations. Based on the DNA-binding domains as annotated in the transcription factor prediction database(Wilson, Charoensawan, Kummerfeld, & Teichmann, 2008), we categorised the 170 analysed TFs into 25 families(Fan et al., 2020). The majority of TFs belonged to the LysR, TetR, AsnC, GntR and AraC families. Among these TFs, PSPPH4700, PSPPH3798, CysB, PSPPH1951, PSPPH4638, PSPPH3504, PSPPH3268 and Irp exhibited over 1,000 binding peaks (Figure S1b). The enriched loci of these binding peaks indicated that these TFs displayed a significant preference for binding to promoters, directly regulating the transcription of downstream targets (Figure S1c). The peak loci of 10 TFs (PSPPH0286, PSPPH0411, PSPPH0711, PSPPH1734, PSPPH2357, PSPPH2407, PSPPH2862, PSPPH3155, PSPPH3431, PSPPH3468, PSPPH4127, PSPPH4622 and PSPPH4768) were completely enriched in the promoter region, with the majority of them belonging to the LysR family. Taken together, the 170 tested TFs in Psph 1448A had over 26,000 DNA-binding peaks distributed across different regions of target genes, suggesting their direct regulatory functions.

Hierarchical TFs reflected multiple higher-order network structures

Transcriptional changes in bacteria are often manipulated by a complex network of TFs. However, bacterial TFs usually have been studied individually or in small clusters with related functions. To comprehensively investigate the associations of all TFs in Psph 1448A at a system level, we constructed a hierarchical network of 270 analysed TFs (100 TFs from HT-SELEX and 170 TFs this study). The findings revealed 1,757 TF interactions among these 26,000 binding events (Supplementary Table 1a). Subsequently, we computed information flow parameters for each TF(Gerstein et al., 2012). In brief, we defined out-degree (O) and in-degree (I) as the number of interactions of a TF in the hierarchical network, representing the regulation of other factors by this TF and the regulation of this TF by other factors, respectively. The difference between O and I indicated the direction of information flow in the network. Hierarchy height (H) was defined as the normalised metric of information circulation, calculated as H = (O − I)/(O + I). When H was close to 1 (H ≈ 1), it indicated that these TFs tended to regulate other factors and occupy upstream positions in the network. Conversely, when H was close to −1 (H ≈ −1), it indicated that these TFs were more likely to be regulated than to regulate other TFs, occupying downstream positions in the network. Based on these criteria, we categorised the 270 analysed TFs (100 TFs from HT-SELEX and 170 TFs this study) into three levels: 54 (20%) executive TFs (such as AlgQ, LexA2 and PSPPH0222) at the top level, 62 (23%) communicative TFs (such as MexT, PsrA and PSPPH1100) at the middle level and 147 (54%) foreman TFs (such as PobR, DksA2 and PSPPH0755) at the bottom level (Figure 1a, Supplementary Table 1b). The presence of a larger number of TFs (147) at the bottom level indicated a high degree of information flow, suggesting the maximisation of the number of downward-pointing edges in the network.

Hierarchical height and collaboration of TFs reveal the multiple regulatory patterns in Psph 1448A. a, Close-up representation of 262 TFs hierarchy in Psph 1448A (eight TFs showed no hierarchical characteristic). Nodes depict TFs. Colors of edges represent source-bases. b, Enrichment of different collaborating (direct interaction, indirect interaction and cooperativity) TF-pairs at top (T), middle (M) and bottom (B) levels. We defined indirect interaction if two TFs co-associated with one target DNA. Cooperative TF pair was defined if their common target is from a TF. Gray nodes below the graph represent TFs. The dashed orange line indicates the averaged level of collaboration. c, Thirteen Three-node sub-modules with the number of occurrences and an example. Spire loop is the most enriched sub-module. Edges represent the regulatory direction. d, Autoregulations are accompanied by the number of occurrences and 13 auto-regulators as examples.

Different numbers of TFs in three levels in the hierarchical network revealed that more than half of TFs (bottom-levels TFs) tend to be regulated by other TFs and then directly bound to target genes. This tendency indicated a downward information flow of transcription regulation of P. syringae. Therefore, we defined the direct binding between two TFs as a direct interaction and investigated collaborations within and between hierarchy levels, specifically intra-level (top to top ‘TT’, middle to middle ‘MM’ and bottom to bottom ‘BB’) and inter-level (top to middle ‘TM’, top to bottom ‘TB’ and middle to bottom ‘MB’) interactions, indirect interaction if two TFs co-associated with one target DNA and cooperative TF pair was defined if their common target was from a TF (Figure 1b, Supplementary Table 1a). In terms of the top-level TFs, direct interactions became more enriched as the hierarchy level of their collaborators decreased. Direct interactions between TB pairs constituted the most substantial portion, accounting for nearly half of all interactions. A similar pattern was observed among the bottom-level TFs, where interactions diminished as the hierarchy level of their collaborators decreased. Compared with interactions among the top- and bottom-level TFs, middle-level TFs, serving as information transmission centres, exhibited lower levels of intra-level collaborations. In summary, transcriptional regulation within the TF hierarchy was predominantly manipulated by top-level TFs, which directed the flow of information to downstream TFs.

Multiple three-node submodules revealed the regulatory diversity of TFs in the Psph 1448A regulatory network

Natural networks, including transcriptional regulation networks, usually show complex characteristics(Newman, 2001; Strogatz, 2001). Among complex networks, some small-scale networks demonstrate numerous connections between individual information nodes and information clusters(Amaral, Scala, Barthelemy, & Stanley, 2000; Jeong, Tombor, Albert, Oltvai, & Barabási, 2000). To investigate the basic structural features of our transcription network, we defined directed edges as direct interactions between two TF nodes and identified global submodules comprising different TF nodes. In this study, we specifically focused on three-node modules, which were considered as ‘network motifs’(Milo et al., 2002). Using algorithms designed to detect recurring modules(Shen-Orr et al., 2002), we scanned our hierarchy network and identified 40,307 different pairs across 13 basic three-node submodules (Figure 1c).

In the first six submodules, we observed that two TF nodes established a relationship only through another node. We denoted these submodules (M1 to M6) as ‘ringent loops’. These seemingly simpler regulation modules appeared more in the Psph 1448A transcriptional regulatory network, especially the first module (M1, n = 24,479), indicating that Psph 1448A favours the use of simple but efficient modes for modulating transcriptional regulation. For example, PhnF and FruR were directly regulated by CysB (M1), and CapA was coregulated by Anr and PSPPH4700 simultaneously (M3, n = 6,177). In addition, M6 (n = 20) contained pairs of mutually regulating TFs (toggle switches), such as PSPPH3268, PSPPH2315 and PSPPH2733.

The remaining seven submodules, denoted as ‘locked loops’ (M7 to M13), comprised subordinate three-node regulatory modules within our network. Notably, no instances of a ‘fully connected triad’ (M12) were observed in our network (n = 0). We found 50 ‘self-loop’ submodules (M7) in our network, including CspD, PSPPH4694 and PSPPH3798, which engaged in mutual regulatory interactions. Among these submodules, M9 (n = 10) was the least common and contained six pairs of toggle switches involving mutually regulating TFs (such as PSPPH1951, PSPPH4612 and CysB), which were similar to those found in the human transcriptional regulatory network(Gerstein et al., 2012). Notably, the most enriched locked loop in our network was M13 (n = 696), denoted as a ‘feed-forward loop’, which has been extensively studied in other species such as humans and E. coli. In this submodule, upstream TFs regulated targeted TFs either by binding directly or manipulating other TFs (Figure S2). For example, TF Irp directly controlled PSPPH4899 and also indirectly regulated it by binding to Lon.

Taken together, the simplest and most effective submodule M1 and the coregulatory submodule M13 played crucial roles in the transcriptional regulation of TFs in Psph 1448A. In addition, we found 92 auto-regulators in our hierarchy network. These auto-regulators are important and always act as repressors in scenarios of multi-stability, such as plant intercellular spaces (Figure 1d) (Alon, 2007). For example, LexA and CysB as negative autoregulators were indicated to reduce cell-to-cell fluctuations in the steady-state level of the transcription factor (Becskei & Serrano, 2000; Rosenfeld, Elowitz, & Alon, 2002). These regulators are regarded as bistable switches that further influence the expression of downstream genes(Burda, Krzywicki, Martin, & Zagorski, 2011). For example, our previous study demonstrated that Lon is a dual-function regulator involved in the regulation of virulence and metabolism in Psph 1448A (Hua et al., 2020). Lon was identified as an auto-regulator in this study. Furthermore, DksA2, which is widely regarded as a protective protein against oxidative stress(Fortuna et al., 2022), was identified as a new bistable switch in this study. In summary, the classification of the TF hierarchy and the identification of enriched network modules not only offer functional predictions for transcriptional regulators but also provide insights into the communication network that governs TF regulation in Psph 1448A.

High co-association pairs occurred more in bottom-level TFs in Psph 1448A

In addition to direct interactions between TFs, we found a notable preference for co-binding peaks among different TF pairs. Briefly, we counted overlapping regions within the binding peaks of all TF pairs. The ratio of intersection regions to the union set of all peaks between two TFs was identified as the genome-wide co-association of specific TFs. We first analysed co-association scores between TF pairs and grouped the scores into 3 TF pairs (TT, MM and BB). The majority of TFs tended to cooperate with other TFs and co-bind to specific genomic regions (Figures S3a, S4a and S5a). The dark distribution in BB pair indicated that high co-association scores preferred to occur in bottom-level TFs. To identify the potential functions of TFs in each level, we used the target genes that were co-bound by TF pairs to perform functional annotations using hypergeometric tests (BH-adjusted pl<l0.05) based on gene sets derived from the Gene Ontology (GO) and Kyoto Encyclopaedia of Genes and Genomes (KEGG) databases (Figures S3b, S4b and S5b Supplementary Table 2a, c–d). The functional categories of TFs at these three levels encompassed various regulatory pathways. For example, the top-level TF PSPPH4700 was involved in siderophore transportation and phosphorelay signal transmembrane transportation. The middle-level TF Lon regulated GTP binding and ribosome transcription. The bottom-level TF PSPPH3486 was involved in amino acid transportation, PSPPH0101 in ATP binding and PSPPH1049 in catalytic activity. Notably, we found that TFs at the top level, without cooperating TFs, exhibited a large number of binding peaks (Figure S3a). This finding suggested that these TFs preferred to regulate target genes by recruiting to specific sites of other TFs, facilitating the direct binding between other TFs and their specific targets, a phenomenon defined as tethered binding (Jie Wang et al., 2012). For example, the top-level TF PSPPH4700 yielded over 1,700 peaks but cooperated with only 24 top-level TFs with low co-association scores about 0.05 (Supplementary Table 2b).

To further elucidate this pattern, we examined the 125 TFs that were analysed through ChIP-seq and exhibited high co-association scores. We determined the co-association patterns among these regulators (Figure 2a) and classified them into four clusters, denoted as C1 to C4. Notably, C1 and C4 contained higher proportions of bottom-level TFs. Based on the analysis in Figure 1b, we found that the proportions of bottom-level TF interaction in all the TF pair interactions and direct interaction were 43% and 49%. These results indicated that the bottom-level TFs tended to regulate downstream genes through cooperating with other level TFs. When comparing the co-association scores of TF pairs, we observed a stronger tendency toward cooperation among lower-level TFs, especially bottom-level TFs. In particular, 35 bottom-level TFs in C4 (83%) exhibited coregulation at the same peak locations with high co-association scores. For instance, PhnF and PSPPH4692 co-bound to three target genes (PSPPH4117, PSPPH4216 and PSPPHB0021), with co-association scores as high as 0.92. This co-binding relationship between all TF pairs was defined as an indirect interaction. In contrast to the trend observed for direct interactions and cooperativity, we observed stronger correlations between top-level TFs and TFs situated at higher levels (top and middle levels; Figure 1b). For example, the top-level TFs PSPPH3798 and CysB co-bound to the promoter region of flhB, which encodes a flagellar biosynthesis protein. We found a similar co-association pattern among bottom-level TFs. However, middle-level TFs displayed the weakest correlations internally, indicating that regulation by higher-level TFs was distributed across different regions in targeted promoters. For instance, we observed that PSPPH3798 bound upstream of flhB, whereas the peak of CysB was located in the overlapStart region.

Bottom TFs share the same binding sequences to coregulate in Psph 1448A. a, The co-association map for 170 TFs in Psph 1448A shows the co-associated scores of binding peaks of these TFs (rows) that overlap each TF peak (columns). The three colored rectangles represent three different TF levels. C1-C4 represent four clusters of TFs according to the co-associated scores. The TFs in corresponding cluster are shown in the circle diagram. Orange nodes represent top TFs. Green nodes represent middle TFs. Purple nodes represent bottom TFs. The colors of edges are the mixture of two source TF colors. b, The heat-map of MexT indicates the associated scores of binding peaks of TFs in C3 (columns) that overlap the binding peaks (rows) of MexT. PSPPH2411, PSPPH3643 and CysI are the top 3 TFs with high associated scores. c, UpSet plot shows the number of genes uniquely targeted TFs or co-targeted by multiple TFs in C3. The vertical black lines represent shared TF-binding sites. The y axis represents the number of overlapped binding sites across the linked TFs. The x axis represents the number of binding sites for each TF. Orange line represents the most enriched gene cysI that is co-targeted by 12 TFs in C3. d, Motifs of MexT and other 10 TFs in C3 which show similar binding sequences.

We found that the correlation between bottom-level TFs was weaker in C3 than in the other clusters. To further explore the DNA-binding characteristics of all TFs within the same cluster, we investigated the peak features of the LysR-family TF MexT in C3 as an example. The peak locations in target genes of seven top-level TFs (PSPPH1435, PSPPH1758, PSPPH2193, PSPPH2454, PSPPH4638, PSPPH4998 and PSPPH3411), three middle-level TFs (PSPPH1100, PSPPH5132 and PSPPH5144) and four bottom-level TFs (PSPPH0700, PSPPH2300, PSPPH2444 and PSPPH2580) were compared with MexT. MexT showed higher co-association scores (more than 60 scores) with more top-level-TFs. The analysis of the peak locations of MexT demonstrated that MexT showed closer co-association relationships with top-level TFs (Figure 2b). Within the binding peaks of MexT, the target genes PSPPH3643 (LysR family TF), PSPPH2618 (sulphite reductase (NADPH) haemoprotein, CysI) and PSPPH2411 (hypothetical protein) displayed high co-association scores with other TFs. UpSet plots demonstrated that 12 TFs in C3 bound to a highly overlapping region of CysI (Figure 2c). These results suggest that these genes were prioritised for co-regulation by MexT. To further explore the binding features in C3, we determined the motifs of 11 TFs based on their binding sequences obtained via ChIP-seq using MEME(Bailey et al., 2009), even though the binding motif of MexT was previously investigated(Tian et al., 2009). We identified a 15-bp consensus motif (ATN11AT) throughout the 11 analysed TFs (Figure 2d), demonstrating a high degree of consensus in co-association patterns among TFs in Psph 1448A.

Virulence-associated pathways were primarily regulated by top-level TFs in Psph 1448A

Because the pathogenicity of P. syringae mainly depends on T3SS and other virulence-associated pathways(He, 1998), we particularly focused on TFs that bind to numerous virulence-related genes. Seven pathways manipulate the virulence of Psph 1448A (Huang et al., 2022; Xie, Shao, & Deng, 2019), namely T3SS, biofilm production, motility, nucleotide-based secondary messenger function, quorum sensing (QS), phytotoxin production and siderophore production. To comprehensively investigate the transcription regulatory mechanism underlying the virulence of Psph 1448A, we calculated the hierarchical heights of TFs involved in virulence regulation and the virulence genes modulated by them. This analysis provided insights into the organisation of the virulence regulatory network in Psph 1448A, where virulence-involved TFs were categorised into three tiers (Figure 3a).

Virulence hierarchical regulatory network reveals 35 TFs involved in virulence. a,

Virulence hierarchical regulatory network shows the TF hierarchy and the large pool of target genes of multi-TF. Target genes are related with seven key virulence pathways, including biofilm formation, secondary messengers, motility, T3SS, QS, phytotoxin production and siderophore transporter. Orange nodes represent top TFs. Green nodes represent middle TFs. Purple nodes represent bottom TFs. Blue nodes represent target genes. Yellow edges represent downward point. Red edges represent upward point. b, The head-to-head binding motif of PSPPH1951, the validation of the binding sites of PSPPH1951 by EMSA, and the detection of expression of target gene hopAE1 in WT, ΔPSPPH1951 and complementary strain by RT-qPCR. The validated binding sites are from the promoters of the hrpR, hopAE1 and hopAH2. c, The binding motif of PSPPH2193 is head-to-head. EMSA confirms the direct binding of PSPPH2193 to the promoters of fleQ and flhF. RT-qPCR confirms the positive regulation of PSPPH2193 on the expression of fleQ and flhF. Motility assay validates the weaker motility of ΔPSPPH2193 than WT and complementary strain. d, The binding motif of PSPPH3268 is head-to-head. EMSA confirms the direct binding of PSPPH3268 to the promoters of hrpR, alg44 and pilM. RT-qPCR confirms the negative regulation of PSPPH3268 on the expression of alg44, algX and pilM. Crystal violate staining assay and the quantification of biofilm formation validate the negative regulation of PSPPH3268 on the biofilm formation. Congo red assay confirms the negative regulation of PSPPH3268 on colony morphologies and EPS production. Student’s t test. n.s., not significant, *PL≤L0.05, **PL≤L0.01, and ***PL≤L0.001.

We found three transcriptional regulatory channels governing virulence regulation in Psph 1448A. The first channel was the direct trigger, which has been extensively studied in previous studies and is referred to as the ‘one-step trigger’ here(Fan et al., 2020). These TFs are recognised as master regulators that directly respond to biological events without additional intermediaries(Chan & Kyba, 2013). In our previous study, we identified TrpI, RhpR, GacA and PSPPH3618 as master regulators in T3SS and 16 master regulators in other virulence pathways(Fan et al., 2020). In line with this definition, we recognised 35 TFs (PSPPH4644, AlgQ, PSPPH1100, CysB, PSPPH2555, PSPPH0239, PSPPH1762, Irp, PSPPH1951, PSPPH2193, PSPH3504, PSPPH1435, PSPPH2983, PSPPH0700, PSPPH5132, PSPPH1776, PSPPH4700, SfsA, PSPPH3522, PSPPH3268, PSPPH4356, PSPPH3798, PSPPH4638, PSPPH4127, Lon, PSPPH4324, PSPPH4673, MexT, PSPPH4844, PSPPH2214, PSPPH0755, PSPPH4012, PSPPH4920, Anr and PSPPH2476) that participate in various virulence pathways. More than 68% of these TFs (24 of 35 TFs) were at the top levels, indicating that the virulence of Psph 1448A is primarily regulated by top-level TFs. Among these TFs in the network, the top-level TFs PSPPH1951, PSPPH2193 and PSPPH3268 were found to have abundant virulence-associated target genes based on ChIP-seq results. The de novo motif analysis of their peak sequences revealed the presence of three head-to-head motifs: a 17-bp motif (AT-N13-AT) for PSPPH1951, a 15-bp motif (ATC-N9-GAT) for PSPPH2193 and a 10-bp motif (AC-N6-GT) for PSPPH3268 (Figure 3b–d, Figure S6a–c).

Three master TFs were identified to participate in virulence

To further verify the biological functions of these three uncharacterised TFs, we first purified the TF proteins and performed an electrophoretic mobility shift assay (EMSA) to confirm their direct interactions with key virulence genes in vitro. Next, we generated TF deletion strains to detect the transcription levels of target genes. We found that PSPPH1951 directly regulated multiple T3SS genes, including hrpRhopAE1 and hopAH2 (Figure 3b). Among them, the expression of hopAE1 in ΔPSPPH1951 was significantly increased by more than four-fold compared with the WT, suggesting that PSPPH1951 acts as a repressor of Psph 1448A T3SS (Figure 3b). In addition, PSPPH1951 was found to bind to the promoters of type IV pili genes (pilG, pilF and pilZ; Figure S6d).

Pst DC3000 enhances its ability to infect the host by increasing bacterial motility(Buell et al., 2003). ChIP-seq data (in vivo) and EMSA (in vitro) results showed that PSPPH2193 interacted with the promoters of motility-related genes, such as fleQ (encoding a flagellar regulator) and flhF (encoding a flagellar biosynthesis regulator; Figure 3c). RT-Qpcr results indicated that fleQ and flhF were downregulated two-fold in ΔPSPPH2193 compared with the WT. As expected, ΔPSPPH2193 exhibited weaker motility than the WT and complementary strain in King’s B medium (Figure 3c), indicating that PSPPH2193 serves as an activator of motility in Psph 1448A.

PSPPH3268 was found to influence the pathogenicity of Psph 1448A by regulating multiple virulence-related pathways. During the initial stage of P. syringae infection, the bacteria produce biofilm components, including extracellular polysaccharides (EPSs), type IV pili and other highly viscous compounds. These components help the bacteria to establish colonies, providing protection against the host’s immune defences and antimicrobial agents(Whitchurch, Tolker-Nielsen, Ragas, & Mattick, 2002). We found that PSPPH3268 strongly interacted with the promoters of key genes involved in biofilm production such as hrpR, the alginate biosynthesis gene alg44 and the type IV pilus assembly gene pilM (Figure 3d). Furthermore, the transcription levels of alg44, algX (encoding the alginate biosynthesis protein) and pilM were markedly enhanced in ΔPSPPH3268. This resulted in enhanced biofilm formation and EPS production when the PSPPH3268 gene was deleted and then restored when PSPPH3268 was expressed (Figure 3d). These results demonstrated that PSPPH3268 acts as a master regulator in various virulence-related pathways. In addition, we identified that the TF PSPPH3798 binds to the promoters of flagellar-related genes (fliK, fliE, fliD and fleQ; Figure S6e).

In addition to the ‘one-step trigger’ mechanism, we found that TFs also regulate downstream genes through one or two other TFs at different levels, which were regarded as ‘one jump-point trigger’ and ‘two jump-point trigger’. For example, PSPPH2555 indirectly influenced biofilm formation (algD), motility (fleQ), T3SS (hrpR), QS (ahlR and secE) and phytotoxin production (aprD) by directly regulating the bottom-level TF PSPPH4920 (Supplementary Table 3). PSPPH3504 was found to be involved in a PSPPH3504-Lon/PSPPH4324/PSPPH4844-PSPPH0755/PSPPH4920/PSPPH4012-tar get gene pathway (Figure S6f, Supplementary Table 3, ‘/’ represents sibling nodes and ‘-’ represents downward regulation). Among these TFs, PSPPH0755, PSPPH4920 and PSPPH4012 are considered key performer TFs because they mediate most of the transcription regulatory signals from multiple TFs. We also found reverse regulatory pathways in our network. For example, the middle-level TF PSPPH4673 was found to directly regulate the top-level TF PSPPH4700 and then indirectly control the transcription of many virulence-related genes through regulating the bottom-level TFs PSPPH0755, PSPPH4920, Anr, and PSPPH2476 (Supplementary Table 3). To further investigate the influence of these TFs on the pathogenicity of P. syringae, we performed the plant infection assay. As expected, we found that the bacterial numbers of Δ2193 strain were significantly reduced compared with WT strain (Figure S6g). In summary, TFs regulate the pathogenicity of Psph 1448A through diverse pathways, either by directly binding to target genes or indirectly controlling other TFs.

Systematic mapping of TF targets revealed key metabolic regulators in Psph 1448A

In addition to enhancing pathogenicity and resisting host defences, P. syringae adjusts its metabolic activities to survive in unpredictable environments(Rico, McCraw, & Preston, 2011). To comprehensively understand metabolic regulation in Psph 1448A, we constructed a hierarchical network that includes key regulators and the genes they trigger, similar to the virulence hierarchical network. We focused on eight metabolic pathways, namely amino acid biosynthesis, DNA replication, ATP-binding cassette (ABC) transportation, oxidative phosphorylation, tricarboxylic acid (TCA) cycle, RNA polymerase, phosphonate metabolism and methyl-accepting chemotaxis (Figure 4a). Compared with the virulence network shown in Figure 3A, we identified more TFs involved in metabolic regulation, many of which exhibited numerous interactions with genes related to oxidative phosphorylation (178 binding peaks) and the TCA cycle (154 binding peaks; Figure S7a–c). In a previous study, 12 master regulators were reported to control various metabolic pathways, including LexA1, PSPPH3004 and PSPPH1960, involved in reactive oxygen species (ROS) resistance(Fan et al., 2020). RhpR participates in several metabolic pathways, such as ABC transporters and oxidoreductase activity(Shao et al., 2019). Lon is involved in glucokinase and oxidoreductase activity(Hua et al., 2020). MgrA, GacA, PilR, PsrA, RpoN, CvsR, OmpR and CbrB2 participate in oxidation resistance, amino acid transportation and other metabolic pathways(Shao et al., 2021). Here, we identified 111 TFs regulating these eight metabolic pathways. Similar to the aforementioned virulence network, three transcriptional regulatory channels were observed for the metabolic pathways. To provide a detailed view of metabolic regulation in Psph 1448A, we counted the number of functionally annotated genes related to each pathway and calculated the proportion of targets for each TF, highlighting key regulators in these eight metabolic pathways using radar plots (Figures 4b and S8a).

Hundreds of TFs are identified to participate in metabolic pathways. a,

Metabolic hierarchical regulatory network shows the TF hierarchy and the large pool of target genes of multi-TF. Target genes are related with eight key metabolic pathways, including biosynthesis of amino acids, DNA replication, ABC transporter, oxidative phosphorylation, TCA cycle, RNA polymerase, phosphonate metabolism and methyl-accepting chemotaxis. Orange nodes represent top TFs. Green nodes represent middle TFs. Purple nodes represent bottom TFs. Blue nodes represent target genes. Yellow edges represent the direct interaction. b, Radar plots show the putative key regulators identified in six different metabolic pathways, including TCA cycle, oxidative phosphorylation, methyl-accepting, biosynthesis of amino acids, RNA polymerase and ABC transporter. Each radiation line represents a key regulator, and the radial length of the thick colored line is the rate of target genes to the associated genes, representing the significance of the enrichment of the TF target genes within each pathway. c, TFs involved in the methyl-accepting pathway bound to the promoters of TFs in the same category. Brown nodes represent the TFs that are responsible for methyl-accepting pathway. The brown arrows point to the targeted TFs. d, TFs involved in the oxidative phosphorylation pathway bind to the promoters of TFs in the methyl-accepting pathway. Red nodes represent the TFs that are responsible for oxidative phosphorylation pathway. Brown nodes represent the TFs that are responsible for methyl-accepting pathway. Blue nodes represent the TFs involved in these two pathways. The arrows point to the targeted TFs and the arrow colors are source-based.

We found that the TFs CysB and PSPPH3268 regulate all eight metabolic pathways, whereas the TFs PSPPH1951, PSPPH3798, PSPPH3504 and PSPPH4700 were predicted to regulate seven metabolic pathways. Notably, the TF PSPPH0755 was found to bind to the promoters of PSPPH5210 (encoding ATP synthase F0F1 subunit delta) and PSPPH3109 (encoding the NADH dehydrogenase subunit A NuoA). The monomer motif (CTGAA) of PSPPH0755 was identified through MEME analysis. The interactions in these two metabolic pathways were confirmed through EMSA (Figure S8b). The TF PSPPH3798 was predicted to bind to the promoters of genes in two pathways, including PSPPH3881 (encoding the methyl-accepting chemotaxis protein WspA) and PSPPH5119 (encoding the phosphate transport system regulatory protein PhoU). The 15-bp binding motif of PSPPH3798 was determined to have a head-to head orientation (ATCG-N7-CGAT). EMSA results confirmed these interactions (Figure S8c). In addition to the TF PSPPH3798, the TF PSPPH4638 had a binding site in the promoter region of the PSPPH3881 gene. PSPPH4638 was also predicted to interact with PSPPH0550 that encodes phosphoserine phosphatase SerB. The 8-bp monomer motif (ATTTTCAA) of PSPPH4638 was identified, and the binding interactions were confirmed through EMSA (Figure S8d).

In yeast, TFs in a functional category appear to bind to genes in the same category(Simon et al., 2001), and we observed a similar pattern in Psph 1448A. For example, TF Irp, a key regulator in the methyl-accepting pathway, bound to the promoters of PSPPH3798, PSPPH4638 and PSPPH4700, which were also identified as key regulators in the same pathway (Figure 4c). In addition, we found that TFs from different categories often bound to the promoters of TFs responsible for other cellular processes. For example, key regulators controlling oxidative phosphorylation (highlighted in red and blue; PSPPH1951, PSPPH2193, PSPPH2832, PSPPH3268, PSPPH3504, PSPPH3798, PSPPH4638, PSPPH4673 and PSPPH4700) bound to TFs playing key roles in the methyl-accepting pathway (highlighted in brown and blue; PSPPH1951, PSPPH3268, PSPPH3504, PSPPH3798, PSPPH4638, PSPPH4700 and Irp; Figure 4d). These results demonstrate that many regulatory processes are often achieved through coregulation by a series of multifunctional TFs throughout the network, enabling Psph 1448A to coordinate transcriptional regulation processes across multiple cellular processes.

TFs indicated large functional variability across different pathovars in P. syringae

Although TF functions exhibit both inter- and intra-species variability, most previous studies on TFs have focused on the molecular mechanism of a single strain(Galardini et al., 2015). To investigate the regulatory mechanism of TFs across different strains of P. syringae, we selected four model strains: P. syringae pv. Syringae 1448A, P. syringae pv. Tomato DC3000, P. syringae pv. Syringae B728a and P. syringae pv. Actinidiae C48. We used the genome of 1448A as a reference and conducted a homology analysis of 1448A protein sequences with those of the other three strains (Supplementary Table 4). We determined a high proportion of homologous proteins in the three strains (4983 in Pst DC3000, 4982 in Pss B728a and 4984 in Psa C48; Supplementary Table 4). Across the four strains, all 301 annotated TFs were present (Supplementary Table 5). We selected five TFs (Irp, PSPPH2193, PSPPH3122, PSPPH4127 and OmpR) to construct TF-overexpressing strains in Pst DC3000, Pss B728a and Psa C48 before performing ChIP-seq analysis. We identified the binding sites of all these TFs in the four strains and found divergent binding preferences for the same TFs in different strains. Most target genes of each TF in one or two strains were unique. In particular, Irp bound to 19 target genes that were conserved in all four tested strains, including purB (encoding adenylosuccinate lyase), cceA2 (encoding the chemotaxis sensor histidine kinase) and gidA (encoding the tRNA uridine 5-carboxymethylaminomethyl modification protein). Four highly conserved target genes were also found to directly interact with PSPPH4127. Evolutionary analysis of the binding peaks of TFs suggested high binding specificity and varying levels of conservation of these TFs in the tested strains (Figure 5a).

Various conservations are observed in TFs between different P. syringae pathovars. a, Proportion of the TF target genes detected in one, two, three or four tested genomes. c1-c4 represent the conservation of targets in one, two, three and four strains. b-c, Repartition of the total pool of target genes (b) or TF-target interactions (c) in four tested strains. d, Enrichment coverage tracks of ChIP-seq against negative controls for TF Irp with binding sites on the promoter of gidA in all four genomes. e, Enrichment coverage tracks of ChIP-seq against negative controls (gray tracks) for the three TFs (Irp, PSPPH2193 and PSPPH4127) with binding sites on the promoter of rpoD in four tested strains.

In addition to the intersection between Psph 1448A and Psa C48, we observed differences between the target genes of all five TFs (Figure 5b) and the peak locations (TF target interactions; Figure 5c) in these four pathovars. The inconsistency between the number of targets and peaks suggested that some target genes were regulated by at least one different TF in these four strains, which is similar to regulation in P. aeruginosa(Trouillon et al., 2021). To confirm the presence of target genes regulated by the same TF or different TFs in various strains, we compared the peak locations of gidA and rpoD (RNA polymerase sigma factor) as two examples. TF Irp was found to bind to the promoter of gidA in four strains (Figure 5d) and had 19 conserved target genes in these strains (Supplementary Table 4). In addition, PSPPH4127 had four conserved target genes (rpoD, PSPPH_1001, PSPPH_1998 and PSPPH_5016) in all four strains (Figure 5e). These results showed that Irp and PSPPH4127 exhibited higher functional conservation than the other three TFs. We also found that PSPPH2193 in 1448A and Irp in Pst DC3000 bound to the promoter of rpoD (Figure 5e). Differences in the regulation of the same targets by different TFs were also observed in more than 1,500 target genes, suggesting the potential diversity of the transcriptional regulation of TFs in our network.

Furthermore, we performed motif and GO analysis to investigate the functional characteristics of TFs (PSPPH3122 and PSPPH4127) in P. syringae strains (Figures S9-S10). As expected, different motifs and specific functions were found in various TFs in four P. syringae strains. For PSPPH3122, motif (AGACN4GATCAA) and motif (CGGACGN3GATCA) were found in Psph 1448A and Pst DC3000 strains, respectively. Functional enrichment analysis of target genes of PSPPH3122 indicated its relationship with recombinase activity and DNA recombination in 1448A strain, while relating with other functions in other strains, such as structural constituent of ribosome in Pss B728a and Pst DC3000 strains. We also found four different motifs of PSPPH4127 in four P. syringae strains. Functional enrichment analysis showed its relationship with recombinase activity in Psph 1448A strain, and RNA binding, structural constituent of ribosome, translation and ribosome in Pss B728a strain. These results showed a highly functional variability of TFs in P. syringae.

Topological modularity of the transcription regulatory network exhibited various functions in biological processes in Psph 1448A

Complex networks in nature often exhibit topological and/or functional modularity(Dittrich, Klau, Rosenwald, Dandekar, & Müller, 2008; Olesen, Bascompte, Dupont, & Jordano, 2007). To explore the modularity in the TF-binding network, we used a partitioning algorithm (with a resolution of 0.9) to classify network elements into different subsets using Gephi. TFs were the primary nodes, and each edge represented the regulatory flow of TFs to target genes, which were small nodes. Our analysed TFs and their target genes were divided into 16 modules, each represented by a different colour (Figure 6a). Multiple applicated modulatory can be achieved depending on different clustering resolution. In our case, a resolution of 0.9 provided moderate modularity and ensured that each module contained 2.7% to 12.1% of the total elements and exhibited correlations with each other. We found that almost all nodes in the network had connections both within and between modules, indicating that the 16 modules were not isolated and contributed to extensive information flow throughout the network to regulate transcription in Psph 1448A (Figure 6b). Module 12 appeared to play a central role in facilitating large information flows with other modules. Module 15 also exhibited transcriptional information exchanges both between modules and within the same module.

Functional modularized regulatory network in P. syringae exhibits the specific functions of both TFs and their target genes. a, The functional categorical regulation network in Psph 1448A analyzed by Gephi (resolution 0.9). The 16 modules (both TFs and their target genes) are labeled in different colors. TF nodes are shown as corresponding sized circles representing their expression level. Their target genes are shown as corresponding-colored dots in the background. TF-target edges are shown as corresponding-colored lines between nodes. b, Graph diagram indicates the connections between TF and their targets in modules. Module nodes are shown as corresponding-colored circles with size proportional to the number of nodes within. Edge colors are source-based, and edge thicknesses represent the connected quantity between modules. c-d, Functional category enrichment analysis of genes in each module, p < 0.05.

Among the 16 modules, Module 2 was involved in most nodes (443 elements, including 41 TFs and 402 target genes). Module 3 contained the least TFs (181 elements, including two TFs and 179 target genes). To investigate the potential correlation between topological modularity and biological functions, we performed GO term and KEGG pathway enrichment analysis for each module by hypergeometric test (BH-adjusted pl<l0.05). As expected, 15 modules were enriched in specific biological functions (Figure 6c and d). In some modules, hundreds of elements were assessed. For example, genes in Module 2 (443 nodes), Module 12 (327 nodes, including 22 TFs and 318 target genes) and Module 14 (432 nodes, including 3 TFs and 324 target genes) were mainly enriched in the regulation of transcription and DNA binding (Figure 6c). DksA, a TF in Module 2, played a key role in regulating transcription-coupled DNA repair(Meddows, Savory, Grove, Moore, & Lloyd, 2005) and also participated in oxidative phosphorylation, amino sugar and nucleotide sugar metabolism and RNA degradation in E. coli. Twenty TFs (such as CapA, CapB, CysB, FruR and MarR) classified in Module 12 were identified to be involved in transcriptional regulation. The TF PSPPH3798 located in Module 14 was observed to be involved in flagellar assembly, and these interactions were confirmed by EMSA (Figure S4e). The genes in Module 4 were enriched in oxidoreductase activity, and our analysis revealed that the TF PSPPH0755 played a role in regulating the oxidative phosphorylation pathway (Figure 6d). MexT in Module 5, which was previously associated with motility in P. syringae pv. tabaci 6605 (Kawakita et al., 2012), was found to participate in in biofilm formation and QS pathways in our study. In addition, we not only identified the crucial roles of the TFs PSPPH1951 (Module 6), PSPPH2193 (Module 11) and PSPPH3268 (Module 15) in T3SS pathways, bacterial motility and biofilm formation, respectively (Figure 3b–d), but also reported their potential biological functions in aminoacyl-tRNA biosynthesis (PSPPH1951), RNA degradation (PSPPH2193) and catalytic activity (PSPPH3268) (Figure 6d). Our results allowed us to identify the potential regulators of specific pathways and perform functional predictions for hypothetical proteins in Psph 1448A. For instance, PSPPH1503 in Module 15, which encodes a hypothetical protein, was possibly correlated with glycerophospholipid metabolism.

Discussion

Most microbial studies on genome-wide transcriptional regulatory network focus on S. cerevisiae and E. coli, which reveal the principles of architecture and interactions of their regulatory networks. The analysis of the transcription regulatory associations in S. cerevisiae mainly rely on the databases such as YEASTRACT (YEAst Search for Transcriptional Regulators And Consensus Tracking)(Teixeira et al., 2018). In E.coli, relative complete transcriptional regulatory network has been generated through integrating three different data sources (RegulonDB, Ecocyc and TRN-SO)(Ma et al., 2004). However, few study has yet comprehensively evaluated TFs in other prokaryotic species throughout a genome(Ishihama, Shimada, & Yamazaki, 2016). In this study, we successfully generated the most complete transcriptional regulatory network and data source, which profiled the transcriptional regulatory features of both the aforementioned 100 TFs (Fan et al., 2020) and an additional 170 TFs in Psph 1448A through ChIP-seq. By mapping the TF-target hierarchical regulatory networks, we identified several novel master regulators involved in significant biological processes. Furthermore, our evolutionary analysis and assessment of the topological functional modularity of TFs and their respective targets revealed the evolutional conservation and functional diversity of TFs in P. syringae.

Although transcriptional regulatory networks are considered conserved(Perez & Groisman, 2009), many studies reveal highly functional variability of TFs in inter- and intra-species(Galardini et al., 2015). These observed diversities between different strains of the same species mainly result from the expression levels of TFs, contents of target genes, and differences of binding sequences(Trouillon et al., 2021). In our study, we observed large differences of DNA-binding characteristic of TF Irp between the C48 strain and the Pst DC3000 strain. The functional diversity of TFs may arise from the large difference in the contents of target genes and TFs, which are regarded as the main determinant of transcriptional regulatory(van Duin, Krautz, Rennie, & Andersson, 2023), although Irp display high homology in these pathovars.

Collaborations between TFs at higher levels (top and middle) were enriched, a pattern similar to the tendencies observed in human TFs(Gerstein et al., 2012). In particular, TT TF pairs exhibited a greater degree of cooperative gene regulation, whereas TB TF pairs accounted for nearly half of direct interactions within all communications. Furthermore, we observed that both direct physical regulation and cooperative interactions were the least common among MM TF pairs. By contrast, in humans, direct regulation tends to occur between TT or TM TF pairs. Furthermore, interactions between TFs, in any form, within human and yeast transcriptional regulatory networks are more likely to appear between middle TFs, which act as information transfer centres(Bhardwaj, Yan, & Gerstein, 2010). This may be attributed to the higher abundance of bottom-level TFs than higher-level TFs observed in prokaryotic microorganisms, a pattern also found in E. coli (Bhardwaj et al., 2010). This finding indicates that the bottom-level TFs that are more likely to be regulated are evolutionarily preferred in multicellular eukaryotic organisms. However, they found that the deficient bottom-level TFs are more commonly co-associated with other same-level TFs in E. coli. When comparing co-associated TF pairs and the cooperativity of TFs, we observed a distinct and inverse relationship in Psph 1448A compared with yeast or E. coli (Figure 1B). The enriched cooperativity of bottom-level TFs with high co-associated scores indicated that these bottom-level TFs preferred to coregulate target genes by binding to the same peak locations. Notably, seven TFs without correlations with other TFs appeared to independently participate in biological processes. These findings not only shed light on the inherent properties of direct regulation and co-association across various species but also indicate the unique characteristics of Psph 1448A to response to dynamic environmental variations.

The fundamental units of a transcriptional regulatory network are positive and negative loops. For a more comprehensive description, these regulatory units can be classified into six submodules, namely autoregulation, multicomponent loops, feedforward loops, single-input, multi-input and regulator chain(Lee et al., 2002). In yeast, only 10 TFs were found to autoregulate themselves, whereas the majority of regulatory units among 116 TFs in E. coli exhibit autoregulation(Thieffry, Huerta, Pérez-Rueda, & Collado-Vides, 1998). Similarly, we identified 92 autoregulators in our transcriptional regulatory network in P. syringae. Feedforward loops are highly prevalent in eukaryotic transcriptional regulatory networks, such as human and yeast(Gerstein et al., 2012; Lee et al., 2002). The 696 feedforward loops (M13) identified in our study also highlighted this cooperative regulation of TFs in response to small signals in prokaryotic species, such as P. syringae. This enriched submodule is regarded as a temporal switch that provides constant feedback to respond rapidly to signal impacts(Shen-Orr et al., 2002). Multistep regulation assists master regulators in enhancing the initial information flow(Goldbeter & Koshland Jr, 1984). Unlike M3, the most abundant module (868) in the human TF regulatory network, we observed that M1 was the most prevalent module (24,479) in the Psph 1448A transcriptional regulatory network. This finding indicates that Psph 1448A prefers achieving transcriptional regulation through few TFs to ensure rapid transmission and response to environmental signals.

Despite more than 20 years of research, our understanding of the global regulatory network in the plant pathogen P. syringae remains inadequate. Our previous studies reported 16 key virulence-related regulators(Shao et al., 2021), 25 master virulence-related regulators(Fan et al., 2020) and seven global regulators acting as TCSs(Xie et al., 2022). However, the regulatory relationships and functional crosstalk among all TFs in P. syringae remain unknown. In this study, we integrated all available interaction information concerning almost all TFs in Psph 1448A and mapped the first comprehensive transcriptional regulatory network in this plant pathogen (Figure 7). This network offers a global view of the multiple functions of TFs in Psph 1448A. We also identified 35 vital TFs that participate in virulence pathways and 111 key TFs involved in metabolic pathways across the global transcriptional regulatory network in Psph 1448A. This analysis uncovered new functions of previously characterised TFs, such as MexT. In addition to the previously reported mexEF-oprN operon,(Sawada et al., 2018) this study identified fleQ (a flagellar regulator) and shcF (a type III chaperone protein) as the targets of MexT. In addition, we investigated the functional evolution and potential intra-species variability of TFs in P. syringae, demonstrating the functional diversity of TFs among P. syringae species during their long course of evolution. Based on the aforementioned results, we established the Psph 1448A transcriptional regulatory network (PSTRnet) database, which contains detailed binding peak information for all TFs in Psph 1448A (https://jiadhuang0417.shinyapps.io/PSTF_NET/). This database serves as a valuable platform for presenting, searching and downloading regulatory information of transcription in Psph 1448A.

Global transcriptional regulatory network in Psph 1448A. The integrated transcriptional regulatory network reflects the interactions between all TFs classified into 39 families from different DNA binding domain types and target genes annotated from pathway annotations. The targets are shown in 12 pathways with various colors. TF nodes are gray as corresponding sizes representing TF number in the family. Edge colors are target-based.

We observed crosstalk between not only TFs but also various pathways. For example, the TF PSPPH4700 directly regulated fleQ and hrpR and indirectly regulated fleQ and hrpR through the PSPPH4700/PSPPH4324/PSPPH0755 cascade. Feedforward loop modules are usually coherent, meaning that the direct effect of downstream TFs has the same regulatory direction (positive or negative) as the indirect effect of upstream TFs(Shen-Orr et al., 2002). This observation enhanced our understanding of the influence of TFs in the network on their target genes based on the identified effects of other TFs. Furthermore, the TF PSPPH4700 was identified to bind to the promoters of genes involved in seven metabolic pathways, namely the TCA cycle, oxidative phosphorylation, methyl-accepting, amino acid biosynthesis, RNA polymerase, ABC transportation and DNA replication. Our results proved the possibility of crosstalk between different pathways in Psph 1448A.

Taken together, our study provides comprehensive insights into the DNA-binding characteristics and potential regulatory pathways of almost all annotated TFs in Psph 1448A. The global transcriptional regulatory network can not only contribute to the development of novel drugs to combat P. syringae infections but also advance research on the molecular mechanisms of TFs in other pathogens.

Materials and Methods

Bacterial strains, culture media, plasmids, and primers

The bacterial strains, plasmids and primers used in this study are listed in Supplementary Table 6. P. syringae Psph 1448A, Pst DC3000, Pss B728a and Psa C48 strains were grown at 28L°C in King’ B (KB) medium shaking at 220 rpm or KB agar plates(King, Ward, & Raney, 1954). E. coli BL21(DE3) or DH5α strains were grown at 37L°C in Luria-Bertani medium (LB) shaking at 220Lrpm or on LB agar plates. Antibiotics used for P. syringae Psph 1448A, Pst DC3000 and Pss B728a wide type (WT) strains and mutants were rifampicin at 50Lμg/ml; P. syringae Psph 1448A, Pst DC3000 and Pss B728a overexpression strains were rifampicin at 50Lμg/ml and spectinomycin at 50Lμg/ml, C48 overexpression strains were spectinomycin at 50Lμg/ml; P. syringae 1448A strains with pK18mobsacB plasmids for mutant construction were rifampicin at 50Lμg/ml and kanamycin at 50Lμg/ml. Antibiotics used for E. coli with pET28a plasmids for protein purification, and pK18mobsacB plasmids were kanamycin at 50Lμg/ml; E. coli with pHM1 plasmids for overexpression strain construction were spectinomycin at 50Lμg/ml.

Construction of overexpression strains and mutants

Overexpression strains and mutants of P. syringae were constructed as previously described(Kvitko & Collmer, 2011; Shao et al., 2021). In brief, for overexpression strain, the open reading frame (Caillet, Baron, Boni, Caillet-Saguy, & Hajnsdorf) of each TF-coding gene was amplified by PCR from P. syringae genome and cloned into pHM1 plasmid. The ligated fragments were inserted into the digested pHM1 plasmids (HindIII) using ClonExpress MultiS One Step Cloning Kit (Vazyme). The recombinant plasmids were transformed into the P. syringae wild-type strain. The single colonies were confirmed by western blot. For mutants, the upstream (∼1500-bp) and downstream (∼1000-bp) fragments of TF ORF were amplified by PCR from the Psph 1448A genome and digested with XbaI respectively, and then ligated by T4 DNA ligase (Jumper et al.). The ligated fragments were inserted into the digested pK18mobsacB plasmids (XbaI) using ClonExpress MultiS One Step Cloning Kit (Vazyme). The recombinant plasmids were transformed into the Psph 1448A wild-type strain. The single colonies were selected to the sucrose plates, and further screened in two kinds of KB plates (with kanamycin/rifampin and with rifampin) concurrently. The single colonies losing kanamycin resistance were further verified by real-time quantitative PCR (RT-qPCR) to detect the mRNA level of corresponding TF genes.

ChIP-seq analyses

Psph 1448A was used for our main ChIP-seq. As previous description(Blasco et al., 2012), the overexpression strains of corresponding TFs and P. syringae WT with empty pHM1 plasmid was cultured in 30 mL KB medium to OD600=0.6. Bacterial cultures were cross-linked with 1% formaldehyde for 10 min at 28°C and then the reaction was stopped by the addition of 125 mM glycine for 5 min. The centrifugated bacteria were washed twice with Tris Buffer (20 mM Tris-HCl [pH 7.5] and 150 mM NaCl) and washed again with IP Buffer (50 mM HEPES–KOH [pH 7.5], 150 mM NaCl, 1 mM EDTA, 1% Triton X-100, 0.1% sodium deoxycholate, 0.1% SDS, and mini-protease inhibitor cocktail). The centrifugated bacteria were preserved at −80L or continued for the next experiments. The bacteria were subjected with IP Buffer and then sonicated to pull down the DNA fragments (150-300-bp). The supernatant was the DNA-TF-HA tag complex and used as IP samples. IP experiments and control sample were incubated with agarose conjugated anti-HA antibodies (Sigma) in IP Buffer. The complex of DNA-TF-anti-HA agarose was applied to washing, crosslink reversal, and purification(Blasco et al., 2012). The 150-250-bp DNA fragments were selected for library construction. The libraries were sequenced using the HiSeq 2000 system (Illumina). Two biological replications have been performed for all ChIP-seq experiments. ChIP-seq reads were mapped to the P. syringae Psph 1448A genome (NC_005773.3), Pst DC3000 (NC_004578.1), Pss B728a (NC_007005.1) and Psa C48 (NZ_CP032631.1) using Bowtie2 (version 2.3.4.3)(Zhang et al., 2008). Subsequently, binding peaks (q < 0.01) were identified using MACS2 software (version 2.1.0). The enriched loci for each TF were annotated using the R package ChIPpeakAnno (version 3.18.2)(L. J. Zhu et al., 2010). We defined intergenic region before each TF sequence as the promoter region. As pHM1 plasmid takes its own promoter, we amplified the TF-coding sequence and cloned into the plasmid. The TF protein expression was activated by the promoter of plasmid.

Electrophoretic mobility-shift assay

DNA probes were amplified from P. syringae 1448A genome by PCR using primers listed in Supplementary Table 6. The probes (20 ng) were mixed with various concentrations of proteins in 20 μL of gel shift buffer (10 mM Tris-HCl, pH 7.4, 50 mM KCl, 5 mM MgCl2, 10% glycerol). After incubation at room temperature for 20 min, the samples were analyzed by 6% native polyacrylamide gel electrophoresis (90 V for 90 min for sample separation). The gels were subjected to Gel Red dye (Tiangen Biotech) for 5 min and photographed by using a gel imaging system (Bio-Rad). The assay was repeated at least three times with similar results.

RT-qPCR

The RT-qPCR primers used are shown Supplementary Table 6 in the Supplemental Information. The cultured bacteria were grown to OD600=0.6 and the total RNA were purified with Bacteria Total RNA Isolation Kit (Sangon Biotech). The RNA concentrations were measured using a NanoDrop 2000 spectrophotometer (ThermoFisher). cDNA synthesis was performed using HiScript III RT SuperMix (Vazyme, China). RT-qPCR was performed with a SuperReal Premix Plus (SYBR Green) kit (Vazyme, China) according to the manufacturer’s instructions. The reactions used 100 nM primers and were run for 40 cycles at 95°C for 30 s and 95°C for 10 s, and at 60°C for 30 s. The fold change represents relative expression level of mRNA, which can be estimated by the values of 2-(ΔΔCt). The relative expression of target genes in WT was set to 1. All the reactions were conducted with three repeats.

Motility assay

The motility assay was performed based on our previous study(Shao et al., 2021). Swimming plates were KB agar plates containing 0.3% agar (MP Biomedicals, UK) and rifampicin at 25 μg/ml. Overnight bacterial cultures were inoculated on swimming plates as 2 μL aliquots and incubated at room temperature for 3-5 days. Finally, the diameter of motility trace represented the swimming motility of P. syringae strains. Photographs were taken by using the Bio-Rad imaging system. The assay was repeated at least three times with similar results.

Congo red assay

Congo red assay was performed as previous study(Shao et al., 2021). Congo red plates were KB agar plates containing 1.0% agar (MP Biomedicals, UK) and rifampicin at 25 μg/ml. Overnight bacterial cultures were inoculated on Congo red plates as 2 μL aliquots and incubated at 28L. The colony staining was photographed after 5-7 days. The assay was repeated at least three times with similar results.

Biofilm formation assay

Biofilm formation assay was performed as previously describe(Shao et al., 2019). Overnight bacterial cultures were transferred to a sterile 10 mL borosilicate tube containing 2 ml KB medium with rifampicin (25 μg/ml) with the original concentration OD600 = 0.1. The cultures grew at room temperature for 36 hr without shaking. 0.1% crystal violet was used to stain the biofilm adhered to the tube for 20 min without shaking. Tubes were washed for more than three times with distilled deionized water gently and other components on the tube loosely was washed off. The tubes were dry and photographed. The remaining crystal violet was fully dissolved in 1 ml 95% ethanol with constantly shaking and measured its optical density at 590 nm (Biotek microplate reader). The assay was repeated at least three times with similar results.

Plant infection assay

Bean leaves were used for pathogenicity. Plant materials were grown in a greenhouse as described previously(Pan et al., 2020; Yanmei Xiao et al., 2007). Overnight bacterial cultures were diluted to OD600=0.2. The diluted cultures were washed three times by sterile water resuspend with equal volume sterile water, and then continued to diluted 100-fold. The diluted bacterial solution was individually injected into bean leaves form stomata on the underside of a leaves. The plant was continued to grow at 28L. Disease symptoms on bean leaves were measured 5 days after inoculation. Infected leaves (1 cm2) were removed after 5 days growth and homogenized in sterile water. Bacteria were diluted to proper concentration and plated on a KB plate containing rifampicin at 25 µg/ml for bacterial count. The bacterial number represented colony-forming unit (CFU) formed in one leaf disks (1 cm2) with four repeats.

Network and functional enrichment analysis

Network analyses were performed on Gephi 0.10. Functional annotations were retrieved from the Pseudomonas database and GO functional enrichment analyses and KEGG analysis were performed using DAVID v6.8.

Statistical analysis

Two-tailed Student’s t tests were performed using Microsoft Office Excel 2010. *PL<L0.05, **PL<L0.01, and ***PL<L0.001 and results represent meansL±LSD. All experiments were repeated at least twice.

Data Availability

Sequencing data have been deposited and publicly available in Gene Expression Omnibus (GEO) under accession number GSE247395. Source data contain the numerical data used to generate the figures of EMSA. Analysis codes have been deposited as accession https://github.com/dengxinb2315/PS-PATRnet-code. Hierarchical information and functional categories of TFs are available in Supplementary Table 1-3. Evolutionary details are provided in Supplementary Table 4. Primers and strains used in this paper are provided in Supplementary Table 6.

Funding

This study was supported by grants from the National Natural Science Foundation of China (32172358 to X.D.), General Research Funds of Hong Kong (11101722, 11102223, and 11102720 to X.D.), Shenzhen Science and Technology Fund (JCYJ20210324134000002 to X.D.), Theme-based Research Scheme (T11-104/22-R to X.D.), Guangdong Major Project of Basic and Applied Basic Research (2020B0301030005 to X.D.). The funders had no role in study design, data collection, interpretation, or the decision to submit the work for publication.

Supporting information